---
params:
  path_cell_type_csv: str
  path_seurat_rds: str
  path_cell_label_reference: str
  ref_name: str
  path_outdir: str
---

# Annotate single-cell cell type based on Tung et al. 2023

## Setup

```{r}
library(Seurat)
library(magrittr)
library(RANN)
library(ggplot2)
library(dplyr)
```

## Parse params

```{r}
path_cell_type_csv <- params$path_cell_type_csv
path_seurat_rds <- params$path_seurat_rds
path_cell_label_reference <- params$path_cell_label_reference
path_outdir <- params$path_outdir
ref_name <- params$ref_name
```

Create output directory

```{r}
dir.create(path_outdir)
```


Cosmetic cell order

```{r}
cell_order <- c("FuO", "RO", "FuEP", "RP", "FuIP", "Ray", "Vessel", "Fiber")
```



## Import data

```{r}
seurat_obj <- readRDS(path_seurat_rds)
cell_label_reference <- read.csv(path_cell_label_reference, row.names = 1)
```

## Find the nearest reference cell in the integrated space

Extract integrated embedding

```{r}
int_embeddings <- Embeddings(seurat_obj, reduction = "integrated.cca")[, 1:30]
```

Set reference

```{r}
is_ref <- seurat_obj$orig.ident == ref_name
seurat_obj$is_ref <- is_ref
```

Perform NN search

```{r}
ref_coords <- int_embeddings[is_ref, ]
query_coords <- int_embeddings[!is_ref, ]
nn_search <- RANN::nn2(
  data = ref_coords,
  query = query_coords,
  k = 1
)
```

Search result

```{r}
query_nn_df <- data.frame(
  query_cell = rownames(query_coords),
  nearest_ref_cell = rownames(ref_coords)[nn_search$nn.idx[, 1]],
  distance = nn_search$nn.dists[, 1]
)
```

Map ref cell barcode to label

```{r}
query_nn_df["Label"] <- cell_label_reference[
  query_nn_df$nearest_ref_cell %>% gsub("_[0-9]$", "", .),
  "Label"
]
query_nn_df[is.na(query_nn_df["Label"]), "Label"] <- "CR_filtered"
```

Update Metadata

```{r}
nn_label <- rep(NA, Cells(seurat_obj) %>% length)
nn_dist <- rep(NA, Cells(seurat_obj) %>% length)
names(nn_label) <- Cells(seurat_obj)
names(nn_dist) <- Cells(seurat_obj)
nn_label[query_nn_df$query_cell] <- query_nn_df$Label
nn_dist[query_nn_df$query_cell] <- query_nn_df$distance
nn_label[row.names(ref_coords)] <- cell_label_reference[
  row.names(ref_coords) %>% gsub("_[0-9]$", "", .),
  "Label"
]
nn_dist[row.names(ref_coords)] <- 0
nn_label[is.na(nn_label)] <- "CR_filtered"
seurat_obj$cell_label <- factor(nn_label, levels = append(
    cell_order, setdiff(unique(nn_label), cell_order)
  )
)
seurat_obj$dist_to_reference <- nn_dist
```

## Summarize cell lable transfer

Color by individual

```{r}
data_plot <- seurat_obj@meta.data[!is_ref, ]
p <- ggplot(
    data_plot, 
    aes(
      x = cell_label,
      y = dist_to_reference,
      fill = individual
    )
  ) + 
  geom_boxplot(
  ) +
  theme_classic(
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Cell type",
    y = "Distance to reference",
    fill = "Individual"
  )
p
ggsave(
  file.path(path_outdir, "distance_to_reference_pre_filter.png"),
  p,
  width = 12,
  height = 6
)
```

## Filter for each cell type the farthest 5% cells to reference

```{r}
meta <- seurat_obj@meta.data
# calculate threshold per cell label
thresholds <- meta %>%
  filter(!is_ref) %>% 
  group_by(cell_label) %>%
  summarise(
    top_5_cutoff = quantile(dist_to_reference, probs = 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# update metadata
meta_updated <- meta %>%
  mutate(cell_barcode = rownames(.)) %>%
  left_join(thresholds, by = "cell_label") %>%
  mutate(cell_label = if_else(
    !is_ref & !is.na(top_5_cutoff) & dist_to_reference >= top_5_cutoff,
    "unannotated",
    cell_label
  ))

rownames(meta_updated) <- meta_updated$cell_barcode
seurat_obj@meta.data <- meta_updated %>% select(-cell_barcode, -top_5_cutoff)
seurat_obj$cell_label <- factor(
  seurat_obj$cell_label, 
  levels = append(cell_order, setdiff(unique(seurat_obj$cell_label), cell_order))
)
```

## Check cell label after filtration

```{r}
data_plot <- seurat_obj@meta.data[!is_ref, ]
p <- ggplot(
    data_plot, 
    aes(
      x = cell_label,
      y = dist_to_reference,
      fill = individual
    )
  ) + 
  geom_boxplot(
  ) +
  theme_classic(
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Cell type",
    y = "Distance to reference",
    fill = "Individual"
  )
p
ggsave(
  file.path(path_outdir, "distance_to_reference_filtered.png"),
  p,
  width = 12,
  height = 6
)
```

Visualize cell label on UMAP

```{r}
p <- DimPlot(
    seurat_obj,
    reduction = "umap.cca_seed17",
    cols = c(
      "#9467BD",
      "#FF7F0F",
      "#2AA02A",
      "#F8E71C",
      "#8C564B",
      "#E377C2",
      "#1F77B4",
      "#D62728",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4"
    ),
    pt.size = 2,
    group.by = "cell_label",
    split.by = "is_ref"
  ) + labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )
ggsave(
  file.path(path_outdir, "umap_cell_label_ref_and_query.png"),
  p,
  width = 12,
  height = 6,
  units = "in",
  dpi = 1000
)
p
```

```{r}
ncol_plot <- seurat_obj$orig.ident %>% unique %>% length %>% sqrt %>% round
p <- DimPlot(
    seurat_obj,
    reduction = "umap.cca_seed17",
    ncol = ncol_plot,
    cols = c(
      "#9467BD",
      "#FF7F0F",
      "#2AA02A",
      "#F8E71C",
      "#8C564B",
      "#E377C2",
      "#1F77B4",
      "#D62728",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4",
      "#d4d4d4"
    ),
    pt.size = 2,
    group.by = "cell_label",
    split.by = "orig.ident"
  ) + labs(
    x = "UMAP 1",
    y = "UMAP 2"
  )
ggsave(
  file.path(path_outdir, "umap_cell_label_by_sample.png"),
  p,
  width = 1 + 5 * (seurat_obj$orig.ident %>% unique %>% length) %/% ncol_plot,
  height = 2 + 5 * ncol_plot,
  units = "in",
  dpi = 1000
)
p
```

Export Cell label

```{r}
write.csv(
  seurat_obj@meta.data %>% filter(!is_ref) %>% select(-nCount_RNA, -nFeature_RNA),
  path_cell_type_csv,
  quote = FALSE
)
```
