---
params:
  path_mtx_dirs: outputs/MetaCell/aggregate_metacell_vartrix/%s
  path_metacell_metadata: tests/MetaCell/metacell_trajectory_weight/metacell_metadata.csv
  path_seurat_obj: outputs/notebooks-new/Seurat_integration/seurat_integrated.rds
  path_hom_loci: outputs/VariantCalling-DNA/gatk_joint/ptr/hom_ref.vcf
  path_variant_annotation: tests/VariantAnnotation-ptr-multi/variant_annotation.tsv
  path_variant_location: tests/VariantAnnotation-ptr-multi/variant_location.tsv
  path_outdir: tests/MetaCell/DEA_Regression
  DEBUG: FALSE
---

## Setup

```{r Setup}
library(glmmTMB)
library(tidyr)
library(dplyr)
library(magrittr)
library(ggplot2)
library(patchwork)
library(Seurat)
library(Matrix)
library(ComplexUpset)
library(data.table)

# parallel
library(parallel)
# Prevent each parallel worker from spawning additional computational threads
Sys.setenv(
  OMP_NUM_THREADS = 1,
  OPENBLAS_NUM_THREADS = 1,
  MKL_NUM_THREADS = 1,
  VECLIB_MAXIMUM_THREADS = 1,
  NUMEXPR_NUM_THREADS = 1
)


cell_order <- c("FuO", "RO", "FuEP", "RP", "FuIP", "Ray", "Vessel", "Fiber")
lineages <- list(
  "Lineage1" = c("FuO", "FuEP", "FuIP", "Fiber"),
  "Lineage2" = c("FuO", "FuEP", "FuIP", "Vessel"),
  "Lineage3" = c("RO", "RP", "Ray")
)
lineage_order <- c(
  "Fusiform late (Fiber)",
  "Fusiform early (Shared)",
  "Fusiform late (Vessel)",
  "Ray Lineage"
)
palette <- c(
  "FuO" = "#9467BD",
  "RO" = "#FF7F0F",
  "FuEP" = "#2AA02A",
  "RP" = "#F8E71C",
  "FuIP" = "#8C564B",
  "Ray" = "#E377C2",
  "Vessel" = "#1F77B4",
  "Fiber" = "#D62728",
  "Outlier_9" = "#d4d4d4",
  "Outlier_10" = "#d4d4d4",
  "Outlier_11" = "#d4d4d4",
  "unannotated" = "#d4d4d4"
)
```

## Parse Rmd parameters

```{r Parse Rmd parameters}
path_mtx_dirs <- params$path_mtx_dirs
path_metacell_metadata <- params$path_metacell_metadata
path_seurat_obj <- params$path_seurat_obj
path_hom_loci <- params$path_hom_loci
path_variant_annotation <- params$path_variant_annotation
path_variant_location <- params$path_variant_location
path_outdir <- params$path_outdir
DEBUG <- params$DEBUG
```

Create output directory

```{r Create output directory}
dir.create(path_outdir, recursive = TRUE)
```

## Import data

### Read genotype data

```{r Read genotype data}
hom_vcf <- fread(
  path_hom_loci,
  skip = "#CHROM",
  header = TRUE
)
head(hom_vcf)
hom_loci <- paste0(hom_vcf$`#CHROM`, ":", hom_vcf$POS)
head(hom_loci)
```

### Read variant annotation data

```{r Read variant annotation data}
var_anno <- fread(path_variant_annotation)
var_loc <- fread(path_variant_location)
var_anno$INDEX <- paste(var_anno$CHROM, var_anno$POS, sep = ":")
var_loc$INDEX <- paste(var_loc$CHROM, var_loc$POS, sep = ":")
setkey(var_anno, INDEX)
setkey(var_loc, INDEX)
```

### Read metacell metadata

```{r Read metacell metadata}
metacell_metadata <- read.csv(path_metacell_metadata, row.names = 1)
metacell_metadata$cell_label_major %<>% factor(levels = cell_order)
for (col_name in  select(metacell_metadata, where(is.character)) %>% colnames) {
  metacell_metadata[[col_name]] %<>% factor
}
row.names(metacell_metadata) <- paste0(
  metacell_metadata$sample,
  "_mc-",
  metacell_metadata$Metacell
)
str(metacell_metadata)
```

### Read ref/alt matrix

```{r Read ref/alt matrix}
alt_mtx_list <- list()
ref_mtx_list <- list()
for (s in metacell_metadata$sample %>% unique) {
  #
  path_mtx_dir_tmp <- sprintf(path_mtx_dirs, s)
  
  # 
  alt_mtx_tmp <- ReadMtx(
    mtx = file.path(path_mtx_dir_tmp, "alt.mtx"),
    cells = file.path(path_mtx_dir_tmp, "col_idx.txt"),
    cell.column = 1,
    features = file.path(path_mtx_dir_tmp, "row_idx.txt"),
    feature.column = 1,
  ) %>% CreateSeuratObject(
    project = s
  ) %>% RenameCells(
    add.cell.id = s
  )
  ref_mtx_tmp <- ReadMtx(
    mtx = file.path(path_mtx_dir_tmp, "ref.mtx"),
    cells = file.path(path_mtx_dir_tmp, "col_idx.txt"),
    cell.column = 1,
    features = file.path(path_mtx_dir_tmp, "row_idx.txt"),
    feature.column = 1,
  ) %>% CreateSeuratObject(
    project = s
  ) %>% RenameCells(
    add.cell.id = s
  )
  
  #
  alt_mtx_list %<>% append(alt_mtx_tmp)
  ref_mtx_list %<>% append(ref_mtx_tmp)
}

# merge objects
alt_mtx <- merge(
  x = alt_mtx_list[[1]],
  y = alt_mtx_list[-1]
)
ref_mtx <- merge(
  x = ref_mtx_list[[1]],
  y = ref_mtx_list[-1]
)

alt_mtx %<>% JoinLayers()
ref_mtx %<>% JoinLayers()

# subset to homozygous reference loci
dim(alt_mtx)
dim(ref_mtx)
alt_mtx %<>% subset(features = hom_loci)
ref_mtx %<>% subset(features = hom_loci)
dim(alt_mtx)
dim(ref_mtx)

# incorporate metadata
alt_mtx@meta.data %<>% cbind(metacell_metadata[Cells(alt_mtx), ])
ref_mtx@meta.data %<>% cbind(metacell_metadata[Cells(ref_mtx), ])

# cleanup
rm(alt_mtx_list, ref_mtx_list, alt_mtx_tmp, ref_mtx_tmp)

#
alt_mtx; ref_mtx; str(alt_mtx@meta.data)
```

## Extract alt/ref read counts

```{r Extract alt/ref read counts}
alt_count <- LayerData(alt_mtx, assay = "RNA", layer = "counts")
ref_count <- LayerData(ref_mtx, assay = "RNA", layer = "counts")
ttl_count <- alt_count + ref_count
```

## QC

### Read depth \~ N Cells per locus

```{r QC}
nCells <- rowSums((alt_count > 0) | (ref_count > 0))
n_cell_avg_depth <- data.frame(
  "n_cell" = nCells,
  "mean_depth" = rowSums(ttl_count) / nCells
)

#
p_xy <- ggplot(
    n_cell_avg_depth,
    aes(
      x = n_cell,
      y = mean_depth
    )
  ) +
  geom_point(
    size = .1
  ) +
  scale_x_log10(
  ) +
  scale_y_log10(
  ) +
  coord_cartesian(
    xlim = c(1, 2000),
    ylim = c(1, 2000)
  ) +
  theme_bw(
  ) +
  labs(
    x = "N Cells",
    y = "Average Read Depth",
    caption = "Each dot equals one locus"
  )
#
p_dens_x <- ggplot(
    n_cell_avg_depth,
    aes(
      x = n_cell,
    )
  ) +
  geom_histogram(
  ) +
  scale_x_log10(
  ) +
  coord_cartesian(
    xlim = c(1, 2000)
  ) +
  theme_classic(
  ) + 
  labs(
    x = NULL
  )
#
p_dens_y <- ggplot(
    n_cell_avg_depth,
    aes(
      y = mean_depth,
    )
  ) +
  geom_histogram(
  ) +
  scale_y_log10(
  ) +
  coord_cartesian(
    ylim = c(1, 2000)
  ) +
  theme_classic(
  ) +
  labs(
    y = NULL
  )

# 
p_patch <- (
    p_dens_x + plot_spacer() +
    p_xy     + p_dens_y
  ) +
    plot_layout(
      ncol = 2,
      widths = c(5, 1),
      heights = c(1, 5),
      byrow = TRUE
    ) &
    theme(
      plot.margin = margin(0, 0, 0, 0, "in")
    )

ggsave(
  p_patch,
  filename = file.path(path_outdir, "average_read_depth_per_locus.png"),
  width = 9,
  height = 6,
  unit = "in",
  dpi = 1000
)
p_patch
```

## Set cutoff on `N cells` and `average read depth` for performing differential RNA editing

```{r Set cutoff}
cutoff_n_cells <- 100
cutoff_avg_depth <- 10
loci_passing_filter <- n_cell_avg_depth %>% filter(
    (n_cell >= cutoff_n_cells) &
    (mean_depth >= cutoff_avg_depth)
  )
p_xy_line <- p_xy + 
  geom_vline(
    xintercept = 100, 
    color = "red", 
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = 10,
    color = "red",
    linetype = "dotted"
  ) +
  annotate(
    "text", 
    x = 500,
    y = 100,
    label = paste0("N = ", nrow(loci_passing_filter)),
    size = 5, 
    colour = "red"
  )

p_patch_cutoff <- (
    p_dens_x  + plot_spacer() +
    p_xy_line + p_dens_y
  ) +
    plot_layout(
      ncol = 2,
      widths = c(5, 1),
      heights = c(1, 5),
      byrow = TRUE
    ) &
    theme(
      plot.margin = margin(0, 0, 0, 0, "in")
    )

ggsave(
  p_patch_cutoff,
  filename = file.path(path_outdir, "average_read_depth_per_locus_cutoff.png"),
  width = 9,
  height = 6,
  unit = "in",
  dpi = 1000
)

p_patch_cutoff
```

## Test model

### Set testing locus and subset data

```{r Test model}
test_locus <- "Chr01:49108455"

# subset metacells  
keep_metacells <- with(
  alt_mtx@meta.data,
  mcRigor_sc == "trustworthy" &
  is.finite(pseudotime) &
  L1_weight > 0.5
)

# get locus data
dat <- ref_mtx@meta.data[keep_metacells, ] %>%
  mutate(
    # get count data
    alt = LayerData(
      alt_mtx, 
      assay = "RNA", 
      layer = "counts",
      cells = Cells(alt_mtx)[keep_metacells],
      features = test_locus
    ) %>% as.numeric,
    ref = LayerData(
      ref_mtx, 
      assay = "RNA", 
      layer = "counts",
      cells = Cells(ref_mtx)[keep_metacells],
      features = test_locus
    ) %>% as.numeric,
    
    # relevel using normal as reference
    treatment = relevel(factor(treatment), ref = "normal"),
    
    # calculate total depth for filtering
    total = alt + ref
  ) %>%
  filter(
    total > 0
  )

# check if treatment/individual are balanced at this locus
print(paste0("Locus: ", test_locus))
table(dat$treatment)
table(dat$individual)
```

### QC

```{r}
p_occurance <- ggplot(
    dat,
    aes(
      x = pseudotime,
      color = treatment,
      fill = treatment
    )
  ) +
  geom_density(
    alpha = 0.15
  ) +
  theme_classic(
  ) +
  labs(
    x = "Pseudotime",
    y = "Density",
    subtitle = "Number of metacells along the pseudotime trajectory"
  )
p_occurance
```

### Model 0: Editing \~ treatment

Does the overall RNA-editing level differ among treatments, including differences caused by their different developmental-state composition?

```{r}
# fit Y ~ treatement model
m0 <- glmmTMB(
  cbind(alt, ref) ~ treatment + (1 | individual) + (1 | sample),
  family = binomial(link = "logit"),
  data = dat
)
summary(m0)
```

Plot editing \~ treatment boxplot

```{r}
p_edit_treatment <- ggplot(
    dat[sample(nrow(dat), replace = FALSE), ],
    aes(
      x = treatment,
      y = alt / total,
      col = treatment
    )
  ) +
  geom_boxplot(
  ) +
  theme_classic(
  ) +
  labs(
    x = "Treatment",
    y = "Editing level",
    col = NULL
  )

p_edit_treatment
```

### Model 1: Editing \~ treatment + overall pseudotime

At comparable developmental positions, is there still a treatment-associated difference in editing?

```{r}
# fit Y ~ treatment + f(T) model
m1 <- glmmTMB(
  cbind(alt, ref) ~ 
    treatment + 
    splines::ns(pseudotime, df = 3) + 
    (1 | individual) +
    (1 | sample),
  family = binomial(link = "logit"),
  data = dat
)
summary(m1)
```

Plot model 1

```{r}
pred_dat <- expand.grid(
  pseudotime = seq(
    min(dat$pseudotime),
    max(dat$pseudotime),
    length.out = 200
  ),
  treatment = levels(dat$treatment),
  individual = dat$individual[1],
  sample = dat$sample[1]
)

pred <- predict(
  m1,
  newdata = pred_dat,
  type = "response",
  se.fit = TRUE,
  re.form = NA
)

pred_dat$fit <- pred$fit
pred_dat$se  <- pred$se.fit

p_m1 <- ggplot(
    dat,
    aes(
      x = pseudotime,
      y = alt / total,
      color = treatment
    )
  ) +
  geom_point(
    aes(
      size = total
    ),
    alpha = 0.15
  ) +
  geom_line(
    data = pred_dat,
    aes(
      y = fit
    ),
    linewidth = 1
  ) +
  scale_size_continuous(
    range = c(0.2, 2)
  ) +
  theme_classic() +
  labs(
    x = "Pseudotime",
    y = "Editing level",
    size = "Read depth"
  )

p_m1
```

#### Model 0 vs 1

```{r}
anova(m0, m1, test = "Chisq")
```

## Parallelized Model Testing

```{r Parallelized Model Testing}
source("notebooks-new/DEA_Regression.R")
```

```{r Smoke test}
candidate_loci <- row.names(loci_passing_filter)
test_results <- run_loci(
  loci = candidate_loci[1:5],
  verbose = TRUE
)

str(test_results$L1_p_treatment_adj)
# should be numeric

str(test_results$L1_n_obs)
# should be integer

str(test_results$L1_ok_additive)
# should be logical

str(test_results$L1_family)
# should be character
```

```{r Run tests}
candidate_loci <- row.names(loci_passing_filter)
system.time({
  results <- run_loci_parallel(
  loci = candidate_loci,
  n_cores = 4,
  weight_cutoff = 0.5,
  spline_df = 3
  )
})
results <- add_fdr(results)
write.csv(
  results,
  file.path(path_outdir, "results.csv")
)
```

## Visualize test results

### 1. Upset plot

Prepare plot data

```{r Plot Upset}
lineages <- c("L1", "L2", "L3")
model_levels <- c(
  "Not tested",
  "Null",
  "Treatment",
  "Pseudotime",
  "Additive"
)

# Prepare data
data_upset <- results %>% 
  select(
    locus,
    L1_model,
    L2_model,
    L3_model
  )

# Create one Boolean column for every lineage × model combination
for (lineage in lineages) {
  model_col <- paste0(lineage, "_model")
  for (model in model_levels) {
    set_col <- paste0(lineage, "__", model)
    data_upset[[set_col]] <- data_upset[[model_col]] == model
  }
}

# Sets in the order we want them displayed
set_cols <- lapply(
    model_levels,
    function(model_level) {
      paste0(lineages, "__", model_level)
    }
  ) %>% unlist

# Pretty labels
set_labels <- setNames(gsub("__", ": ", set_cols) %>% unlist, set_cols)
```

Plot

```{r}
p_model_upset <- ComplexUpset::upset(
  data_upset,
  intersect = set_cols,
  # Every locus should have exactly:
  # one L1 state + one L2 state + one L3 state
  min_degree = 3,
  max_degree = 3,
  min_size = 1, # Prevent empty patterns from overwhelming the plot
  sort_sets = FALSE, # Keep L1/L2/L3 ordering
  sort_intersections_by = "cardinality", # Largest combinations first
  width_ratio = 0.18,
  encode_sets = FALSE,
  labeller = function(x) {
    sub("__", ": ", x)
  },
  base_annotations = list(
    "Number of loci" = intersection_size(counts = TRUE)
  ),
  set_sizes = upset_set_size(
    geom = geom_bar(width = 0.7)
  )
) +
  labs(
    title = "RNA editing model patterns across developmental lineages"
  )

ggsave(
  p_model_upset,
  filename = file.path(path_outdir, "upset_lineage_model.png"),
  width = 12,
  height = 8,
  unit = "in",
  dpi = 1000
)

p_model_upset
```

### 2. Editing \~ Pseudotime (per locus)

```{r Plot editing ~ Pseudotime}
path_plot_out <- file.path(path_outdir, "editing_levels")
dir.create(path_plot_out, showWarnings = TRUE, recursive = TRUE)
for (row_idx in 1:nrow(results)) {
  locus <- results[row_idx, "locus"] %>% sub(":", "_", .)
  p_editing_time <- plot_locus_result(results[row_idx,])
  ggsave(
    p_editing_time,
    filename = file.path(path_plot_out, paste0(locus, ".png")),
    width = 12,
    height = 8,
    unit = "in",
    dpi = 1000
  ) 
}

```

## Combine with Variant Annotation

```{r Add variant annotation}
# find loci where pseudotime and/or treatment significantly effect editing
loci_any <- results %>%
  select(
    L1_model,
    L2_model,
    L3_model
  ) %>%
  mutate(
    L1_model = L1_model %in% c("Pseudotime", "Treatment", "Additive"),
    L2_model = L2_model %in% c("Pseudotime", "Treatment", "Additive"),
    L3_model = L3_model %in% c("Pseudotime", "Treatment", "Additive")
  )
loci_any <- results[rowSums(loci_any) > 0, "locus"]

# Prevent inflation by ignoring variant annotation on transcript isoforms unless the consequence are different
var_data <- full_join(
  var_anno[intersect(loci_any, var_anno$INDEX)] %>% 
    distinct(
      INDEX, GENEID, CONSEQUENCE,
      .keep_all = TRUE
    ),
  var_loc[loci_any] %>% 
    distinct(
      INDEX, GENEID, LOCATION,
      .keep_all = TRUE
    ),
  by = c("INDEX", "GENEID")
) %>%
  select(
    -CHROM.x,
    -CHROM.y,
    -POS.x,
    -POS.y
  ) %>%
  relocate(
    INDEX
  )

# plot
p_var_anno <- ggplot(
    var_data,
    aes(
      x = LOCATION,
      fill = CONSEQUENCE
    )
  ) +
  geom_bar(
  ) +
  geom_text(
    aes(label = after_stat(count)),
    stat = "count", 
    position = position_stack(vjust = 0.5)
  ) +
  theme_bw(
  ) +
  labs(
    x = "Location on a gene",
    y = "Counts",
    fill = "Outcome"
  )

ggsave(
  p_var_anno,
  filename = file.path(path_outdir, "editing_outcome.png"),
  width = 6,
  height = 4,
  unit = "in",
  dpi = 1000
)

p_var_anno
```

Export outcome of editing

```{r Export editing outcomes}
write.csv(
  var_data,
  file.path(path_outdir, "editing_outcome.csv"),
  row.names = FALSE
)
```
