---
params:
  template_opt_mc: "outputs/MetaCell/mcrigor/{sample}/metacell.csv"
  template_opt_rds: "outputs/MetaCell/mcrigor/{sample}/optimize_res.rds"
  template_path_h5: "outputs/CellRanger/count/{sample}/outs/filtered_feature_bc_matrix.h5"
---

# Summarizing MetaCell & McRigor results

MetaCell analysis was performed on each dataset using varying combinations of K (30, 40, 50, 60, 70) and alpha (0.6, 0.8. 1.0, 1.2) for `metacell::mcell_mc_from_coclust_balanced`. (Script: "notebooks-new/MetaCell.Rmd")

McRigor was used to find the best metacell partitioning parameters for each dataset. (Script: "notebooks-new/McRigor.Rmd")

Both pipelines were called using Snakemake. The rules were in "workflows/MetaCell.smk"

## Setup

Library

```{r}
library(glue)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Seurat)
```

Vars

```{r}
template_opt_mc <- params$template_opt_mc
template_opt_rds <- params$template_opt_rds
template_path_h5 <- params$template_path_h5
sample_names <- c(
  "ptr_tenx_batch2",
  "ptr_tenx_tsv2", 
  "ptr_tenx_tsv3", 
  "ptr_tenx_tsv4", 
  "ptr_tenx_tsv5",
  "ptr_tenx_tst2", 
  "ptr_tenx_tst3", 
  "ptr_tenx_tst4",
  "ptr_tenx_tso2", 
  "ptr_tenx_tso3", 
  "ptr_tenx_tso4"
)
sample_names <- c(
  "ptr_tenx_tso3", 
  "ptr_tenx_tso4"
)
```

Import optimized metacell partitions

```{r}
mc_labels <- data.frame()
for (s in sample_names) {
  foo <- read.csv(glue(template_opt_mc, sample=s))
  names(foo)[names(foo) == "X"] <- "barcode"
  foo["sample"] <- s
  mc_labels %<>% rbind(foo)
}
mc_labels$sample <- factor(mc_labels$sample, levels = sample_names)
summary(mc_labels)
```

Box plot of cell sizes (x = sample; y = sizes; col = trustworthy/dubious)

```{r}
data_plot <- mc_labels %>% group_by(sample, Metacell, mcRigor_sc) %>% count(name = "size")
p_box <- ggplot(
    data = data_plot, 
    aes(x = sample, y = size, fill = mcRigor_sc)
  ) + 
  geom_boxplot(
    width = 0.6,
    position = position_dodge(width = 0.8)
  ) +
  scale_fill_manual(
    values = c(
      "trustworthy" = "#72caed", 
      "dubious" = "#f57d81"
    )
  ) +
  theme_classic(
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    y = "Cell sizes",
    fill = "mcRigor"
  )
p_box
```

Bar graph of MetaCell counts

```{r}
p_bar <- ggplot(
    data = data_plot,
    aes(x = sample, fill = mcRigor_sc)
  ) + geom_bar(
    width = 0.8,
    position="fill"
  ) +
  scale_fill_manual(
    values = c(
      "trustworthy" = "#72caed", 
      "dubious" = "#f57d81"
    )
  ) +
  theme_classic(
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    y = "Percentage",
    fill = "mcRigor"
  )
p_bar
```

Combine plots and output

```{r}
p_patch <- p_box / p_bar
p_patch
```

Plot MetaCells on single-cell UMAP embeddings

```{r}
p_mc_list <- list()
for (s in sample_names) {
  # read mcRigor results
  optimize_res <- readRDS(glue(template_opt_rds, sample = s))
  opt_metacell <- optimize_res$opt_metacell
  sc_membership <- opt_metacell@misc$cell_membership$Metacell
  names(sc_membership) <- rownames(opt_metacell@misc$cell_membership)
  
  # read single-cell
  obj_singlecell <- Read10X_h5(
    glue(template_path_h5, sample=s)
  ) %>% CreateSeuratObject()
  obj_singlecell <- NormalizeData(obj_singlecell)
  obj_singlecell <- FindVariableFeatures(obj_singlecell)
  obj_singlecell <- ScaleData(obj_singlecell)
  obj_singlecell <- RunPCA(obj_singlecell)
  obj_singlecell <- RunUMAP(
    obj_singlecell, 
    dims = 1:30,
    metric = "euclidean",
    n.neighbors = 30,
    min.dist = .1,
    seed.use = 42
  )
  
  break
  p_mc <- mcRigor_projection(
    obj_singlecell = obj_singlecell, 
    sc_membership = sc_membership,
    color_field = 'celltype.l1',
    dub_mc_test.label = T, 
    test_stats = optimize_res$TabMC, 
    Thre = optimize_res$thre
  )
  p_mc_list[[s]] <- p_mc
}

```
