---
params:
  path_metacell_annotated: tests/metacell_annotation/metacell_annotation.csv
  path_mc_label_template: outputs/MetaCell/mcrigor/{sample}/metacell.csv
  path_cell_type_annotation: outputs/notebooks-new/Cell_type_annotation/cell_type_annotation.csv
  path_outdir: outputs/MetaCell/metacell_trajectory_weight
  path_seurat_obj: outputs/notebooks-new/Seurat_integration/seurat_integrated.rds
---

# Annotate Metacell cell label based on single-cell label transferred from [Tung et al.](https://doi.org/10.1186/s13059-022-02845-1)

## Setup

```{r}
library(dplyr)
library(magrittr)
library(glue)
library(ggplot2)
library(patchwork)
library(Seurat)
library(slingshot)

cell_order <- c("FuO", "RO", "FuEP", "RP", "FuIP", "Ray", "Vessel", "Fiber")
palette <- c(
  "#9467BD",
  "#FF7F0F",
  "#2AA02A",
  "#F8E71C",
  "#8C564B",
  "#E377C2",
  "#1F77B4",
  "#D62728",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4"
)
```

## Parse Rmd parameters

```{r}
path_metacell_annotated <- params$path_metacell_annotated
path_mc_label_template <- params$path_mc_label_template
path_cell_type_annotation <- params$path_cell_type_annotation
path_outdir <- params$path_outdir
path_seurat_obj <- params$path_seurat_obj
```

## Create output directory

```{r}
dir.create(path_outdir)
```

## Import data

Read transferred single-cell annoation

```{r}
cell_type_annotation <- read.csv(path_cell_type_annotation, row.names = 1)
cell_type_annotation["barcode"] <- row.names(cell_type_annotation)
cell_type_annotation
```

Read MetaCell membership labels

```{r}
metacell_memberships <- data.frame()
for (sname in unique(cell_type_annotation$orig.ident)) {
  tmp <- read.csv(glue(path_mc_label_template, sample = sname), row.names = 1)
  sample_suffix <- cell_type_annotation %>% 
    subset(orig.ident == sname) %>% 
    row.names %>% 
    head(1) %>% 
    strsplit("_") %>% 
    unlist() %>%
    extract(2)
  rownames(tmp) %<>% paste(sample_suffix, sep = "_")
  metacell_memberships %<>% rbind(tmp)
}
metacell_memberships["barcode"] <- row.names(metacell_memberships)
metacell_memberships
```

## Read Seurat object

```{r}
seurat_obj <- readRDS(path_seurat_obj)
seurat_obj <- subset(
  seurat_obj, 
  seurat_obj$orig.ident %in% unique(cell_type_annotation$orig.ident)
)
```

## Import lineage pseudo-time

```{r}

```

## Merge MetaCell membership & Cell label annotation data frame

```{r}
metacell_annotated <- full_join(
  x = metacell_memberships,
  y = cell_type_annotation,
  by = "barcode"
)
row.names(metacell_annotated) <- metacell_annotated$barcode
metacell_annotated %<>% select(-barcode)
metacell_annotated
```

Filter NA caused by missing values during full_join

(1) Find the columns containing NA

```{r}
col_na <- metacell_annotated %>% is.na %>% apply(2, any) %>% which %>% names
col_na
```

(2) See if NA coincide per observation

```{r}
metacell_annotated[, col_na] %>% is.na %>% apply(1, sum) %>% table
```

(3) Exclude NA-containing rows

```{r}
metacell_annotated %<>% 
  subset(metacell_annotated[, col_na] %>% is.na %>% not %>% apply(1, any))
metacell_annotated
```

## Calculate the purity of cell types per metacell per sample

### 1. Purity Score

This measures the proportion of cells belonging to the most dominant category within a metacell.

-   **Formula**:

    $$Purity = \max_i(p_i) = \frac{N_{max}}{N_{total}}$$

-   **Interpretation**: Higher values indicate a more homogeneous metacell. A value of $1$ indicates that all cells belong to the same category.

-   **Example**: If a metacell contains 10 cells (8 from category A and 2 from category B), the purity is:

    $$Purity = \frac{8}{10} = 0.8$$

Purity is intuitive, but it only considers the abundance of the dominant category and ignores how the remaining cells are distributed among other categories.

------------------------------------------------------------------------

### 2. Normalized Shannon Entropy

Normalized Shannon entropy measures the overall heterogeneity, or impurity, of the category composition within a metacell. Unlike purity, it considers the relative abundance of **all categories**.

First, Shannon entropy is calculated as:

$$H = -\sum_{i=1}^{K} p_i \log_2(p_i)$$

where $p_i$ is the proportion of cells assigned to category $i$.

To make entropy comparable across different annotation schemes, it is normalized by the maximum possible entropy:

$$H_{norm} = \frac{-\sum_{i=1}^{K} p_i \log_2(p_i)}{\log_2(K)}$$

where $K$ is the **fixed total number of possible categories in the dataset** (e.g., the total number of `cell_label` or `cca_clusters` categories), rather than the number of categories observed within an individual metacell.

-   **Interpretation**:

    -   $H_{norm} = 0$: perfectly pure; all cells belong to one category.
    -   $H_{norm} \rightarrow 1$: increasingly heterogeneous.
    -   $H_{norm} = 1$: maximum heterogeneity, where cells are equally distributed among all $K$ possible categories.

-   **Example**: If there are two possible categories and a metacell contains 8 cells from category A and 2 cells from category B:

    $$H = -(0.8\log_2(0.8) + 0.2\log_2(0.2)) \approx 0.72$$

    Because $K = 2$:

    $$H_{norm} = \frac{0.72}{\log_2(2)} = 0.72$$

Thus, higher normalized entropy represents greater metacell impurity, whereas higher purity represents greater metacell homogeneity.

For the metacell analysis, normalized entropy is used as the primary measure of impurity because it incorporates the full distribution of cell identities. Purity is retained as a complementary and readily interpretable measure of the proportion represented by the dominant category.

```{r}
major_category <- function(x) {
  tab <- table(x, useNA = "no")
  names(tab)[which.max(tab)]
}
calc_purity <- function(x) {
  if (length(x) == 0) return(NA)
  tab <- table(x, useNA = "no")
  max(tab) / sum(tab)
}
calc_entropy <- function(x, K) {
  if (length(x) == 0) return(NA)
  tab <- table(x, useNA = "no")
  p <- as.numeric(tab) / sum(tab)
  H <- -sum(p * log(p))
  H / log(K)
}

K_cca   <- n_distinct(metacell_annotated$cca_clusters)
K_label <- n_distinct(metacell_annotated$cell_label)

metacell_summary <- metacell_annotated %>%
  # mcRigor_sc and is_ref are intentionally excluded from the summary
  group_by(orig.ident, Metacell) %>%
  summarise(
    sample = first(orig.ident),
    individual = first(individual),
    treatment  = first(treatment),
    mcRigor_sc = first(mcRigor_sc),

    cca_clusters_major   = major_category(cca_clusters),
    cca_clusters_purity  = calc_purity(cca_clusters),
    cca_clusters_entropy = calc_entropy(cca_clusters, K_cca),

    cell_label_major   = major_category(cell_label),
    cell_label_purity  = calc_purity(cell_label),
    cell_label_entropy = calc_entropy(cell_label, K_label),

    dist_to_reference_mean = mean(dist_to_reference, na.rm = TRUE),
    dist_to_reference_sd   = sd(dist_to_reference, na.rm = TRUE),

    n_cells = n(),

    .groups = "drop"
  ) %>% select(-orig.ident)

# cell_label as factor (with custom levels)
cell_levels <- c(cell_order, setdiff(metacell_summary$cell_label_major, cell_order) %>% sort)
metacell_summary$cell_label_major %<>% factor(levels = cell_levels)
metacell_summary$mcRigor_sc %<>% factor(levels = c("trustworthy", "dubious"))

# convert other character columns to factor
for (col_name in  select(metacell_summary, where(is.character)) %>% colnames) {
  metacell_summary[[col_name]] %<>% factor
}
metacell_summary
```

## Visualization

### SD\~Mean scatter plot of "distance to reference cells". Colors show the purity metrics of cell label.

1.  Purity score

    ```{r}
    p_purity <- ggplot(
        metacell_summary,
        aes(
          x = dist_to_reference_mean,
          y = dist_to_reference_sd,
          col = cell_label_purity
        )
      ) + 
      geom_point(
        size=.5
      ) +
      geom_smooth(
        formula = y ~ x, 
        method = "lm"
      ) + 
      scale_color_viridis_c(
        direction = -1
      ) +
      theme_classic(
      ) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
      labs(
        x = "Average distance to reference cell",
        y = "SD of distance to reference cell",
        col = "Purity score"
      )

    # save figure
    ggsave(
      p_purity,
      filename = file.path(path_outdir, "sd~mean_purity.png"),
      width = 8,
      height = 6,
      unit = "in",
      dpi = 1000
    )

    p_purity
    ```

<!-- -->

2.  Normalized Shannon Entropy

    ```{r}
    p_entropy <- ggplot(
        metacell_summary,
        aes(
          x = dist_to_reference_mean,
          y = dist_to_reference_sd,
          col = cell_label_entropy
        )
      ) + 
      geom_point(
        size=.5
      ) +
      geom_smooth(
        formula = y ~ x, 
        method = "lm"
      ) + 
      scale_color_viridis_c(
      ) +
      theme_classic(
      ) +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
      labs(
        x = "Average distance to reference cell",
        y = "SD of distance to reference cell",
        col = "Normalized\nShannon Entropy"
      )

    # save figure
    ggsave(
      p_entropy,
      filename = file.path(path_outdir, "sd~mean_entropy.png"),
      width = 8,
      height = 6,
      unit = "in",
      dpi = 1000
    )

    p_entropy
    ```

### Numbers of trustworthy/dubious metacells per cell label per treatment

```{r}
p_bar <- ggplot(
    metacell_summary,
    aes(
      x = cell_label_major,
      fill = mcRigor_sc
    )
  ) +
  geom_bar(
    width = 0.6,
    position = position_dodge(0.8, preserve = "single")
  ) +
  theme_linedraw(
  ) + 
  scale_fill_manual(
    values = c(
      "trustworthy" = "#72caed", 
      "dubious" = "#f57d81"
    )
  ) + 
  facet_grid(
    individual ~ treatment
  ) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1)
  ) +
  labs(
    x = "Major cell label",
    y = "Number of metacells",
    fill = "McRigor"
  ) 

# save figure
ggsave(
  p_bar,
  filename = file.path(path_outdir, "number_of_metacells.png"),
  width = 10,
  height = 10,
  unit = "in",
  dpi = 1000
)

# print figure
p_bar
```

### Box plot of cell type purity metrics

1.  Purity score

    ```{r}
    p_box_purity <- ggplot(
        metacell_summary,
        aes(
          x = cell_label_major,
          y = cell_label_purity,
          fill = mcRigor_sc
        )
      ) + 
      geom_boxplot(
        position = position_dodge(0.8, preserve = "single"),
        width = 0.6
      ) +
      scale_fill_manual(
        values = c(
          "trustworthy" = "#72caed", 
          "dubious" = "#f57d81"
        )
      ) + 
      facet_grid(
        individual ~ treatment
      ) +
      theme_linedraw(
      ) + 
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
      labs(
        x = "Major cell label",
        y = "Purity score",
      )

    # save figure
    ggsave(
      p_box_purity,
      filename = file.path(path_outdir, "boxplot_purity.png"),
      width = 10,
      height = 10,
      unit = "in",
      dpi = 1000
    )

    p_box_purity
    ```

2.  Normalized Shannon Entropy

    ```{r}
    p_box_entropy <- ggplot(
        metacell_summary,
        aes(
          x = cell_label_major,
          y = cell_label_entropy,
          fill = mcRigor_sc
        )
      ) + 
      geom_boxplot(
        position = position_dodge(0.8, preserve = "single"),
        width = 0.6
      ) +
      scale_fill_manual(
        values = c(
          "trustworthy" = "#72caed", 
          "dubious" = "#f57d81"
        )
      ) + 
      facet_grid(
        individual ~ treatment
      ) +
      theme_linedraw(
      ) + 
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1)
      ) +
      labs(
        x = "Major cell label",
        y = "Normalized Shannon Entropy",
      )

    # save figure
    ggsave(
      p_box_entropy,
      filename = file.path(path_outdir, "boxplot_entropy.png"),
      width = 10,
      height = 10,
      unit = "in",
      dpi = 1000
    )

    p_box_entropy
    ```

## Filter Metacell with too high normalized entropy

-   McRigor classification indicates whether a metacell is statistically homogeneous

-   Normalized Entropy of cell type labels quantifies whether a metacell is representative of a single discrete cell type

Check if "trustworthy", yet "high entropy" metacells were at the junction of discrete cell types

```{r}
seurat_obj <- subset(
  seurat_obj,
  seurat_obj$orig.ident %in% unique(cell_type_annotation$orig.ident)
)
seurat_obj$cell_label <- cell_type_annotation[Cells(seurat_obj), "cell_label"]
seurat_obj$cell_label %<>% factor(levels = cell_order)

emb <- Embeddings(seurat_obj, "umap.cca_seed17") %>% as.data.frame
colnames(emb) <- c("UMAP_1", "UMAP_2")
emb$Metacell <- metacell_annotated[row.names(emb), "Metacell"]
emb$sample <- metacell_annotated[row.names(emb), "orig.ident"]
metacell_coordinates <- emb %>% 
  group_by(sample, Metacell) %>%
  summarize(
    "UMAP_1" = median(UMAP_1),
    "UMAP_2" = median(UMAP_2),
    .groups = "drop"
  )
metacell_coordinates <- full_join(
  metacell_coordinates,
  metacell_summary,
  by = c("Metacell", "sample")
)
```

```{r}
p_umap <- DimPlot(
    seurat_obj, 
    group.by = "cell_label",
    cols = palette,
    reduction = "umap.cca_seed17"
  ) +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Cell label",
    title = NULL
  )
p_label_entropy <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = cell_label_entropy
    )
   ) + 
  geom_point(
  ) + 
  scale_color_viridis_c(
  ) +
  theme_classic(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Normalized Entropy (Cell label)"
  )
p_mcrigor <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = mcRigor_sc
    )
   ) + 
  geom_point(
  ) + 
  scale_color_manual(
    values = c(
      "trustworthy" = "#72caed", 
      "dubious" = "#f57d81"
    )
  ) +
  theme_classic(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "mcRigor"
  )
p4 <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = cell_label_major
    )
   ) + 
  geom_point(
  ) + 
  scale_color_manual(
    values = palette
  ) +
  theme_classic(
  ) +
  theme(
    legend.position = "none"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Cell label",
    title = "Metacells (Cell label determined via majority voting)"
  )

patch <- (p_umap | p4) / (p_mcrigor | p_label_entropy)

ggsave(
  patch,
  filename = file.path(path_outdir, "umap_overlay.png"),
  width = 12,
  height = 12,
  unit = "in",
  dpi = 1000
)

patch
```
