---
params:
  sample_name: ptr_tenx_tsv2
  alpha: 2
  K: 100
  path_mcdb: str
  path_figs: str
  path_10x: str
  path_output: str
  path_mc_label: str
---

# ```{r, setup, include=FALSE}
# # set this option in the first code chunk in the document
# knitr::opts_chunk$set(
#   echo = params$sample_name,
#   root.dir = "/home/rstudio/local"
# )
# ```

# Metacell analysis

Library metacell

```{r Library metacell}
library(metacell)
library(magrittr)
library(R.utils)
```

## Setup

```{r Setup}
sample_name <- params$sample_name
alpha <- params$alpha %>% as.numeric
K <- params$K %>% as.numeric
path_mcdb <- params$path_mcdb
path_figs <- params$path_figs
path_10x <- params$path_10x
path_output <- params$path_output
path_mc_label <- params$path_mc_label
```

Initiate MetaCell DB

```{r scdb_init}
dir.create(path_mcdb, recursive = TRUE)
scdb_init(path_mcdb, force_reinit = TRUE)
```

Load 10x matrix

```{r mcell_import_scmat_10x}
temp_files <- sapply(
  c(
    file.path(path_10x, "barcodes.tsv.gz"),
    file.path(path_10x, "features.tsv.gz"), 
    file.path(path_10x, "matrix.mtx.gz")
  ),
  FUN = gunzip,
  remove = FALSE,
  temporary = TRUE
)
mcell_import_scmat_10x(
  mat_nm = sample_name,
  base_dir = dirname(temp_files)[1]
)
sapply(
  temp_files,
  FUN = file.remove
)
mat = scdb_mat(sample_name)
print(dim(mat@mat))
```

Initiate MetaCell fig directory

```{r scfigs_init}
dir.create(path_figs, recursive = TRUE)
scfigs_init(path_figs)
```

Show UMI distribution

```{r mcell_plot_umis_per_cell}
mcell_plot_umis_per_cell(sample_name)
```

## Selecting feature genes

```{r mcell_add_gene_stat}
mcell_add_gene_stat(
  gstat_id = sample_name, 
  mat_id = sample_name, 
  force = T
)
```

Create a gset object

```{r mcell_gset_filter_varmean}
mcell_gset_filter_varmean(
  gset_id = sprintf("%s_feats", sample_name), 
  gstat_id = sample_name, 
  T_vm = 0.2, 
  force_new = T
)
mcell_gset_filter_cov(
  gset_id = sprintf("%s_feats", sample_name), 
  gstat_id = sample_name, 
  T_tot = 100, 
  T_top3 = 2
)
```

Variance-mean plot

```{r mcell_plot_gstats}
mcell_plot_gstats(
  gstat_id = sample_name, 
  gset_id = sprintf("%s_feats", sample_name)
)
```

## Building the balanced cell graph

```{r mcell_add_cgraph_from_mat_bknn}
mcell_add_cgraph_from_mat_bknn(
  mat_id = sample_name, 
  gset_id = sprintf("%s_feats", sample_name), 
  graph_id = sprintf("%s_graph", sample_name),
  K = 100,
  dsamp = T
)
```

## Resampling and generating the co-clustering graph

```{r mcell_coclust_from_graph_resamp}
mcell_coclust_from_graph_resamp(
  coc_id = sprintf("%s_coc500", sample_name), 
  graph_id = sprintf("%s_graph", sample_name),
  min_mc_size = 20, 
  p_resamp = 0.75, 
  n_resamp = 500
)
```

```{r mcell_mc_from_coclust_balanced}
mcell_mc_from_coclust_balanced(
  coc_id = sprintf("%s_coc500", sample_name), 
  mat_id = sample_name,
  mc_id = sprintf("%s_mc", sample_name), 
  K = K, 
  min_mc_size = 30,
  alpha = alpha
)
```

## Removing outlier cells

Skipped this step and check for dubious MetaCells using mcRigor

```{r mcell_mc_split_filt}
mcell_mc_split_filt(
  new_mc_id = sprintf("%s_mc_f", sample_name),
  mc_id = sprintf("%s_mc", sample_name),
  mat_id = sample_name,
  T_lfc = 3,
  plot_mats = F
)
```

## Selecting markers and coloring metacells

```{r mcell_gset_from_mc_markers}
mcell_gset_from_mc_markers(
   gset_id = sprintf("%s_markers", sample_name),
   mc_id = sprintf("%s_mc", sample_name)
)
```

Plot marker gene heatmap

```{r mcell_mc_plot_marks}
mcell_mc_plot_marks(
  mc_id = sprintf("%s_mc", sample_name),
  gset_id =sprintf("%s_markers", sample_name),
  mat_id = sample_name
)
mc_colorize_default(sprintf("%s_mc", sample_name))
```

Plot MetaCell graph

```{r mcell_mc2d_force_knn}
mcell_mc2d_force_knn(
  mc2d_id = sprintf("%s_2dproj", sample_name),
  mc_id = sprintf("%s_mc", sample_name),
  graph_id = sprintf("%s_graph", sample_name)
)
tgconfig::set_param("mcell_mc2d_height", 1000, "metacell")
tgconfig::set_param("mcell_mc2d_width", 1000, "metacell")
mcell_mc2d_plot(mc2d_id=sprintf("%s_2dproj", sample_name))
```

## Export MetaCell label

```{r export}
mc <- scdb_mc(sprintf("%s_mc_f", sample_name))
mc_label <- mc@mc %>% as.data.frame
colnames(mc_label) <- as.character(K)
write.csv(mc_label, path_mc_label)
```
