---
params:
  path_metacell_annotated: str
  path_mc_label_template: str
  path_cell_type_annotation: str
  path_pseudo_time: str
  path_lineage_weight: str
  path_lineage_prob: str
  path_branch_id: str
  path_seurat_obj: str
  path_outdir: str
  path_metacell_metadata: str
  DEBUG: bool
---

# Aggregate cell label and order per metacell

Pseudotime was inferred using Slingshot (2.4.0) on CCA-integrated single-cell embeddings. Single-cell cell labels were transferred from nearest [Tung et al.](https://doi.org/10.1186/s13059-022-02845-1) bio1 neighbors in the same space.

## Setup

```{r Setup}
library(dplyr)
library(magrittr)
library(ggplot2)
library(patchwork)
library(Seurat)

cell_order <- c("FuO", "RO", "FuEP", "RP", "FuIP", "Ray", "Vessel", "Fiber")
lineages <- list(
  "Lineage1" = c("FuO", "FuEP", "FuIP", "Fiber"),
  "Lineage2" = c("FuO", "FuEP", "FuIP", "Vessel"),
  "Lineage3" = c("RO", "RP", "Ray")
)
lineage_order <- c(
  "Fusiform late (Fiber)",
  "Fusiform early (Shared)",
  "Fusiform late (Vessel)",
  "Ray Lineage"
)
palette <- c(
  "FuO" = "#9467BD",
  "RO" = "#FF7F0F",
  "FuEP" = "#2AA02A",
  "RP" = "#F8E71C",
  "FuIP" = "#8C564B",
  "Ray" = "#E377C2",
  "Vessel" = "#1F77B4",
  "Fiber" = "#D62728",
  "Outlier_9" = "#d4d4d4",
  "Outlier_10" = "#d4d4d4",
  "Outlier_11" = "#d4d4d4",
  "unannotated" = "#d4d4d4"
)
```

## Parse Rmd parameters

```{r Parse Rmd parameters}
path_metacell_annotated <- params$path_metacell_annotated
path_mc_label_template <- params$path_mc_label_template
path_cell_type_annotation <- params$path_cell_type_annotation
path_pseudo_time <- params$path_pseudo_time
path_lineage_weight <- params$path_lineage_weight
path_lineage_prob <- params$path_lineage_prob
path_branch_id <- params$path_branch_id
path_seurat_obj <- params$path_seurat_obj
path_outdir <- params$path_outdir
path_metacell_metadata <- params$path_metacell_metadata
DEBUG <- as.logical(params$DEBUG)
```

## Create output directory

```{r dir.create}
dir.create(path_outdir, recursive = TRUE)
```

## Import data

Read discrete cell label

```{r Read discrete cell label}
cell_type_annotation <- read.csv(path_cell_type_annotation, row.names = 1)
cell_type_annotation["barcode"] <- row.names(cell_type_annotation)
str(cell_type_annotation)
```

Read Slingshot output matrices (pseudotime/lineage weight/normalized lineage probability/branch id)

```{r Read Slingshot output matrices}
mtx_pseudotime <- read.csv(path_pseudo_time, row.names = 1)
mtx_weight <- read.csv(path_lineage_weight, row.names = 1)
mtx_prob <- read.csv(path_lineage_prob, row.names = 1)
mtx_branch_id <- read.csv(path_branch_id, row.names = 1)

colnames(mtx_pseudotime) <- paste0(colnames(mtx_pseudotime), "_pseudotime")
colnames(mtx_weight) <- paste0(colnames(mtx_weight), "_weight")
colnames(mtx_prob) <- paste0(colnames(mtx_prob), "_prob")

mtx_pseudotime["barcode"] <- row.names(mtx_pseudotime)
mtx_weight["barcode"] <- row.names(mtx_weight)
mtx_prob["barcode"] <- row.names(mtx_prob)
mtx_branch_id["barcode"] <- row.names(mtx_branch_id)

mtx_slingshot <- mtx_pseudotime %>% 
  full_join(mtx_weight, by = "barcode") %>% 
  full_join(mtx_prob, by = "barcode") %>% 
  full_join(mtx_branch_id, by = "barcode")
str(mtx_slingshot)
```

Read MetaCell membership labels

```{r Read MetaCell membership labels}
metacell_memberships <- data.frame()
for (sname in unique(cell_type_annotation$orig.ident)) {
  tmp <- read.csv(sprintf(path_mc_label_template, sname), row.names = 1)
  sample_suffix <- cell_type_annotation %>% 
    subset(orig.ident == sname) %>% 
    row.names %>% 
    head(1) %>% 
    strsplit("_") %>% 
    unlist() %>%
    extract(2)
  rownames(tmp) %<>% paste(sample_suffix, sep = "_")
  metacell_memberships %<>% rbind(tmp)
}
metacell_memberships["barcode"] <- row.names(metacell_memberships)
str(metacell_memberships)
```

## Read Seurat object

```{r Read Seurat object}
seurat_obj <- readRDS(path_seurat_obj)
seurat_obj <- subset(
  seurat_obj,
  orig.ident %in% unique(cell_type_annotation$orig.ident)
)
```

## Merge metacell, cell label, and slingshot data frame

```{r Merge MetaCell membership & Cell label annotation}
metacell_annotated <- metacell_memberships %>% 
  full_join(cell_type_annotation, by = "barcode") %>% 
  full_join(mtx_slingshot, by = "barcode") 
row.names(metacell_annotated) <- metacell_annotated$barcode
metacell_annotated %<>% mutate(
  L12_shared = pmin(Lineage1_weight, Lineage2_weight),
  L1_specific = pmax(Lineage1_weight - Lineage2_weight, 0),
  L2_specific = pmax(Lineage2_weight - Lineage1_weight, 0)
)
metacell_annotated %<>%
  mutate(
    branch_id = case_match(
      branch_id,
      "1" ~ "Fusiform late (Fiber)", 
      "2" ~ "Fusiform late (Vessel)",
      "1,2" ~ "Fusiform early (Shared)",
      "3" ~ "Ray Lineage",
      "" ~ "Unknown",
     .default = branch_id
    )
  )
if (DEBUG) {metacell_annotated <- metacell_annotated[mtx_slingshot$barcode, ]}
str(metacell_annotated)
```

Filter NA (cells missing in MetaCell output)

```{r Filter cells missing in MetaCell output}
metacell_annotated %<>% subset(Metacell %>% is.na %>% not)
str(metacell_annotated)
```

## Aggregate pseudotime and lineage weight per metacell

For each metacell $m$ and lineage $l$:

-   **Lineage weight**:

    $$
    W_{m,l} = \frac{1}{N_m}\sum_{c \in m} w_{c,l}
    $$

    where $N_m$ is the total number of cells in metacell $m$, and $w_{c,l}$ is the lineage weight of cell $c$ for lineage $l$.

-   **Lineage-specific pseudotime**:

    $$
    T_{m,l} =
    \frac{
    \sum_{c \in m} w_{c,l} t_{c,l}
    }{
    \sum_{c \in m} w_{c,l}
    }
    $$

    where $t_{c,l}$ is the pseudotime of cell $c$ along lineage $l$.

    Equivalently, because

```{r Aggregate data per metacell}
major_category <- function(x) {
  tab <- table(x, useNA = "no")
  names(tab)[which.max(tab)]
}
calc_purity <- function(x) {
  if (length(x) == 0) return(NA)
  tab <- table(x, useNA = "no")
  max(tab) / sum(tab)
}
calc_entropy <- function(x, K) {
  if (length(x) == 0) return(NA)
  tab <- table(x, useNA = "no")
  p <- as.numeric(tab) / sum(tab)
  H <- -sum(p * log(p))
  H / log(K)
}
weighted_sd <- function(pt, w) {
  keep <- is.finite(pt) & is.finite(w) & w > 0
  pt <- pt[keep]
  w  <- w[keep]
  if (length(pt) < 2 || sum(w) == 0)
      return(NA_real_)
  mu <- weighted.mean(pt, w)
  sqrt(
      sum(w * (pt - mu)^2) /
    sum(w)
  )
}

K_cca   <- n_distinct(metacell_annotated$cca_clusters)
K_label <- n_distinct(metacell_annotated$cell_label)

metacell_summary <- metacell_annotated %>%
  # mcRigor_sc and is_ref are intentionally excluded from the summary
  group_by(orig.ident, Metacell) %>%
  summarise(
    sample = first(orig.ident),
    individual = first(individual),
    treatment  = first(treatment),
    mcRigor_sc = first(mcRigor_sc),

    cca_clusters_major   = major_category(cca_clusters),
    cca_clusters_purity  = calc_purity(cca_clusters),
    cca_clusters_entropy = calc_entropy(cca_clusters, K_cca),

    cell_label_major   = major_category(cell_label),
    cell_label_purity  = calc_purity(cell_label),
    cell_label_entropy = calc_entropy(cell_label, K_label),

    dist_to_reference_mean = mean(dist_to_reference, na.rm = TRUE),
    dist_to_reference_sd   = sd(dist_to_reference, na.rm = TRUE),
    
    L1_pseudotime = weighted.mean(Lineage1_pseudotime, w = Lineage1_weight),
    L2_pseudotime = weighted.mean(Lineage2_pseudotime, w = Lineage2_weight),
    L3_pseudotime = weighted.mean(Lineage3_pseudotime, w = Lineage3_weight),
    L1_weight = mean(Lineage1_weight, na.rm = TRUE),
    L2_weight = mean(Lineage2_weight, na.rm = TRUE),
    L3_weight = mean(Lineage3_weight, na.rm = TRUE),
    L1_prob = mean(Lineage1_prob, na.rm = TRUE),
    L2_prob = mean(Lineage2_prob, na.rm = TRUE),
    L3_prob = mean(Lineage3_prob, na.rm = TRUE),
    L1_sd = weighted_sd(Lineage1_pseudotime, w = Lineage1_weight),
    L2_sd = weighted_sd(Lineage2_pseudotime, w = Lineage2_weight),
    L3_sd = weighted_sd(Lineage3_pseudotime, w = Lineage3_weight),
    
    L12_shared   = mean(L12_shared),
    L1_specific  = mean(L1_specific),
    L2_specific  = mean(L2_specific),
    
    branch_major = major_category(branch_id),
    n_cells = n(),

    .groups = "drop"
  ) %>% select(-orig.ident)

# aggregate pseudotime across lineages
weighted_pt <- function(pt, prob) {
  keep <- is.finite(pt) & is.finite(prob) & prob > 0

  if (!any(keep))
    return(NA_real_)

  prob <- prob[keep] / sum(prob[keep])

  sum(pt[keep] * prob)
}
mixture_sd <- function(pt, sd, prob) {
  keep <- is.finite(pt) &
          is.finite(sd) &
          is.finite(prob) &
          prob > 0

  if (!any(keep))
    return(NA_real_)

  pt   <- pt[keep]
  sd   <- sd[keep]
  prob <- prob[keep] / sum(prob[keep])

  mu <- sum(prob * pt)

  sqrt(
    sum(
      prob * (sd^2 + (pt - mu)^2)
    )
  )
}
metacell_summary <- metacell_summary %>%
  rowwise() %>%
  mutate(
    pseudotime = weighted_pt(
      c(L1_pseudotime, L2_pseudotime, L3_pseudotime),
      c(L1_prob, L2_prob, L3_prob)
    ),
    SD = mixture_sd(
      c(L1_pseudotime, L2_pseudotime, L3_pseudotime),
      c(L1_sd, L2_sd, L3_sd),
      c(L1_prob, L2_prob, L3_prob)
    )
  ) %>%
  ungroup()


# cell_label as factor (with custom levels)
cell_levels <- c(cell_order, setdiff(metacell_summary$cell_label_major, cell_order) %>% sort)
metacell_summary$cell_label_major %<>% factor(levels = cell_levels)
metacell_summary$mcRigor_sc %<>% factor(levels = c("trustworthy", "dubious"))

# convert other character columns to factor
for (col_name in  select(metacell_summary, where(is.character)) %>% colnames) {
  metacell_summary[[col_name]] %<>% factor
}
str(metacell_summary)
```

## Visualization

### Overlay statistics on integrated UMAP embedding

Get metacell coordinates

```{r Get metacell coordinates}
# add cell label to metadata
seurat_obj$cell_label <- cell_type_annotation[Cells(seurat_obj), "cell_label"]
seurat_obj$cell_label %<>% factor(levels = cell_order)

# add slingshot output to metadata
if (DEBUG) {seurat_obj %<>% subset(cells = mtx_slingshot$barcode)}
seurat_obj@meta.data %<>% cbind(mtx_slingshot)
seurat_obj@meta.data %<>% mutate (
    pseudotime = (
      Lineage1_pseudotime * Lineage1_weight +         
      Lineage2_pseudotime * Lineage2_weight +
      Lineage3_pseudotime * Lineage3_weight
    ) / (Lineage1_weight + Lineage2_weight + Lineage3_weight)
  )

# calculate metacell coordinates
emb <- Embeddings(seurat_obj, "umap.cca_seed17") %>% as.data.frame
colnames(emb) <- c("UMAP_1", "UMAP_2")
emb$Metacell <- metacell_annotated[row.names(emb), "Metacell"]
emb$sample <- metacell_annotated[row.names(emb), "orig.ident"]
metacell_coordinates <- emb %>% 
  group_by(sample, Metacell) %>%
  summarize(
    "UMAP_1" = median(UMAP_1),
    "UMAP_2" = median(UMAP_2),
    .groups = "drop"
  )
metacell_coordinates <- full_join(
  metacell_coordinates,
  metacell_summary,
  by = c("Metacell", "sample")
)
```

Single-cell label, Metacell label, mcRigor, Entropy

```{r plot UMAP overlay 1}
# single cell cell label 
p_sc_labeled <- DimPlot(
    seurat_obj, 
    group.by = "cell_label",
    cols = palette,
    reduction = "umap.cca_seed17"
  ) + 
  theme_void(
  ) +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Cell label",
    axis.text = element_blank(),
    title = NULL
  )

# metacell cell label entropy
p_label_entropy <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = cell_label_entropy
    )
   ) + 
  geom_point(
  ) + 
  scale_color_viridis_c(
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Normalized Entropy (Cell label)"
  )

# metacell trustworthiness (mcRigor)
p_mcrigor <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = mcRigor_sc
    )
   ) + 
  geom_point(
  ) + 
  scale_color_manual(
    values = c(
      "trustworthy" = "#72caed", 
      "dubious" = "#f57d81"
    )
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "mcRigor"
  )

# metacell cell label
p_mc_labeled <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = cell_label_major
    )
   ) + 
  geom_point(
  ) + 
  scale_color_manual(
    values = palette
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "none"
  ) +
  labs(
    x = NULL,
    y = NULL
  )

patch <- (p_sc_labeled | p_mc_labeled) / (p_mcrigor | p_label_entropy)

ggsave(
  patch,
  filename = file.path(path_outdir, "umap_overlay.png"),
  width = 12,
  height = 12,
  unit = "in",
  dpi = 1000
)

patch
```

```{r plot UMAP overlay 2}
# single cell cell label 
p_sc_l1 <- FeaturePlot(
    seurat_obj, 
    features = c("pseudotime"),
    reduction = "umap.cca_seed17"
  ) + 
  scale_color_viridis_c(
    direction = -1,
    option = "magma"
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "top"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    col = "Pseudotime",
    axis.text = element_blank(),
    title = NULL
  )

# metacell lineage 1
p_mc_time <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = pseudotime
    )
   ) + 
  geom_point(
  ) + 
  scale_color_viridis_c(
    direction = -1,
    option = "magma"
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = NULL,
    y = NULL,
    col = "Pseudotime"
  )

# metacell pseudotime SD
p_mc_sd <- ggplot(
    data = metacell_coordinates,
    aes(
      x = UMAP_1,
      y = UMAP_2,
      col = SD
    )
   ) + 
  geom_point(
  ) + 
  scale_color_viridis_c(
    option = "viridis"
  ) +
  theme_void(
  ) +
  theme(
    legend.position = "bottom"
  ) +
  labs(
    x = NULL,
    y = NULL,
    col = "Pseudotime SD",
  )

patchwork <- ( p_sc_labeled | p_sc_l1 ) / ( p_mc_time | p_mc_sd )

ggsave(
  patchwork,
  filename = file.path(path_outdir, "umap_overlay2.png"),
  width = 12,
  height = 12,
  unit = "in",
  dpi = 1000
)

patchwork
```

### Metacell pseudotime (with SD) by lineages

```{r plot metacell pseudotime + SD}
data_plot <- metacell_summary
data_plot$branch_major %<>% factor(levels = lineage_order)
data_plot %<>% filter(branch_major %>% is.na %>% not)

#
p_branch_time <- ggplot(
    data_plot,
    aes(
      x = pseudotime, 
      y = branch_major,
      col = cell_label_major
    )
  ) +
  geom_point(
    size = 2
  ) +
  scale_color_manual(
    values = palette
  ) +
  facet_grid(
    mcRigor_sc~.
  ) +
  theme_linedraw(
  ) +
  theme(
    legend.position = "right"
  ) +
  labs(
    x = "Average Pseudotime",
    y = "Lineages",
    col = "Major cell label",
  )

#
p_time_sd <- ggplot(
    data_plot,
    aes(
      x = pseudotime, 
      y = pseudotime,
      ymax = pseudotime + SD,
      ymin = pseudotime - SD,
      col = cell_label_major
    )
  ) +
  geom_pointrange(
    linewidth = .2,
    size = .02
  ) +
  scale_color_manual(
    values = palette
  ) +
  facet_grid(
    branch_major~mcRigor_sc
  ) +
  theme_linedraw(
  ) +
  theme(
    legend.position = "right"
  ) +
  labs(
    x = "Average Pseudotime",
    y = "Average Pseudotime",
    col = "Major cell label",
    subtitle = "dots: average pseudotime; errorbar: ±sd"
  )

ggsave(
  p_branch_time,
  filename = file.path(path_outdir, "pseudotime_by_branch.png"),
  width = 12,
  height = 6,
  unit = "in",
  dpi = 1000
)
ggsave(
  p_time_sd,
  filename = file.path(path_outdir, "pseudotime_sd.png"),
  width = 12,
  height = 5,
  unit = "in",
  dpi = 1000
)

p_branch_time; p_time_sd
```

## Final output data.frame

### Pseudotime on Fusiform Lineage (with branching topology)

```{r}
data_plot <- metacell_summary %>%
  mutate(
    fusiform_prob = L1_prob + L2_prob,
    branch_position = if_else(
      fusiform_prob > 0,
      (L2_prob - L1_prob) / fusiform_prob,
      NA_real_
    ),
    fusiform_pseudotime = if_else(
      fusiform_prob > 0,
      (L1_pseudotime * L1_prob + L2_pseudotime * L2_prob) / fusiform_prob,
      NA_real_
    )
  )

p_fusiform <- ggplot(
    data_plot %>% filter(fusiform_prob > 0),
    aes(
      x = fusiform_pseudotime,
      y = branch_position,
      color = cell_label_major
    )
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  geom_point(
    aes(alpha = fusiform_prob),
    size = 2
  ) +
  scale_y_continuous(
    limits = c(-1.05, 1.05),
    breaks = c(-1, 0, 1),
    labels = c(
      "Fiber",
      "Fusiform shared",
      "Vessel"
    )
  ) +
  scale_color_manual(
    values = palette
  ) +
  scale_alpha_continuous(
    range = c(.01, 1)
  ) +
  labs(
    x = "Fusiform pseudotime",
    y = "Branch position",
    color = "Major cell label",
    alpha = "Fusiform\nprobability"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank()
  )

ggsave(
  p_fusiform,
  filename = file.path(path_outdir, "pseudotime_fusiform.png"),
  width = 9,
  height = 5,
  unit = "in",
  dpi = 1000
)

p_fusiform
```

```{r}
p_ray <- ggplot(
    metacell_summary %>% filter(L3_prob > 0),
    aes(
      x = L3_pseudotime,
      y = L3_prob,
      color = cell_label_major,
      alpha = L3_prob
    )
  ) +
  geom_point(
    size = 2
  ) +
  scale_y_continuous(
    limits = c(-.05, 1.05),
    breaks = c(0, 1),
    labels = c("Fusiform", "Ray")
  ) +
  scale_color_manual(
    values = palette
  ) +
  labs(
    x = "Ray pseudotime",
    y = "Branch position",
    color = "Major cell label",
    alpha = "Ray\nprobability"
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

ggsave(
  p_ray,
  filename = file.path(path_outdir, "pseudotime_ray.png"),
  width = 9,
  height = 4,
  unit = "in",
  dpi = 1000
)

p_ray
```

## Export Metacell metadata

```{r}
write.csv(
  metacell_summary,
  path_metacell_metadata
)
```
