# Integrating ***P. trichocarpa*** **samples using Seurat**

## Setup

Library

```{r}
library(Seurat)
library(magrittr)
library(ggplot2)
library(dplyr)
```

Global variables

```{r}
samples <- c(
  "ptr_tenx_batch1", "ptr_tenx_batch2",
  "ptr_tenx_tsv1", "ptr_tenx_tsv2", "ptr_tenx_tsv3", "ptr_tenx_tsv4", "ptr_tenx_tsv5",
  "ptr_tenx_tst1", "ptr_tenx_tst2", "ptr_tenx_tst3", "ptr_tenx_tst4",
  "ptr_tenx_tso1", "ptr_tenx_tso2", "ptr_tenx_tso3", "ptr_tenx_tso4"
)
treatment <- c(
  "normal", "normal",
  "normal", "normal", "normal", "normal", "normal",
  "tension", "tension", "tension", "tension",
  "opposite", "opposite", "opposite", "opposite"
)
individual <- c(
  "tung_1", "tung_2",
  "hsieh_1", "hsieh_2", "hsieh_3", "hsieh_4", "hsieh_5",
  "hsieh_6", "hsieh_7", "hsieh_8", "hsieh_9",
  "hsieh_6", "hsieh_7", "hsieh_8", "hsieh_9"
)
metadata <- data.frame(
  sample = samples,
  treatment = treatment,
  individual = individual
)
rownames(metadata) <- samples
head(metadata)

# 10x data path
path_h5 <- "outputs/CellRanger/count/{sample}/outs/filtered_feature_bc_matrix.h5"
```

## Read 10x data

Merged into one SeuratObject with separated layers for each sample

```{r}
obj_list <- lapply(
  samples, function(sname) {
    filename <- sub("[{]sample[}]", sname, path_h5)
    obj <- Read10X_h5(filename) %>% CreateSeuratObject(project = sname)
    return(obj)
  }
)
seurat_obj <- merge(
  x = obj_list[[1]],
  y = obj_list[-1]
)
rm(obj_list)
seurat_obj
```

Add metadata to SeuratObject

```{r}
seurat_obj$individual <- metadata[seurat_obj$orig.ident, "individual"]
seurat_obj$treatment <- metadata[seurat_obj$orig.ident, "treatment"]
```

Plot UMI counts and Feature counts per sample

```{r}
p_counts <- VlnPlot(
  seurat_obj, 
  features = "nCount_RNA",
  group.by = "orig.ident",
  pt.size = .05
) + NoLegend()
p_features <- VlnPlot(
  seurat_obj, 
  features = "nFeature_RNA", 
  group.by = "orig.ident",
  pt.size = .05
) + NoLegend()
p_counts | p_features
```

### Exclude samples with low UMI counts

```{r}
excluded_samples <- c(
  "ptr_tenx_batch1",
  "ptr_tenx_tso1",
  "ptr_tenx_tst1",
  "ptr_tenx_tsv1"
)
seurat_obj <- subset(
  seurat_obj,
  idents = excluded_samples, 
  invert = TRUE
)
print(seurat_obj)
metadata <- subset(
  metadata, 
  !rownames(metadata) %in% excluded_samples
)
print(metadata)
```

## Pre-processing

```{r}
seurat_obj <- NormalizeData(seurat_obj)
seurat_obj <- FindVariableFeatures(seurat_obj)
seurat_obj <- ScaleData(seurat_obj)
seurat_obj <- RunPCA(seurat_obj)
```

```{r}
ElbowPlot(seurat_obj, 50)
```

## UMAP (pre-integration)

```{r}
seurat_obj <- RunUMAP(
  seurat_obj, 
  dims = 1:30,
  metric = "euclidean",
  n.neighbors = 30,
  min.dist = .1,
  seed.use = 49
)
p_individual <- DimPlot(seurat_obj, group.by = "individual")
p_treatment <- DimPlot(seurat_obj, group.by = "treatment")
p_individual | p_treatment
```

## Integration

```{r}
seurat_obj <- IntegrateLayers(
  seurat_obj,
  method = CCAIntegration,
  dims = 1:20,
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  sample.tree = matrix(c(
    -1, -2,
    1, -3,
    2, -4,
    3, -5,
    -6, -7,
    5, -8,
    -9, -10,
    7, -11,
    4, 6,
    9, 8
  ),
  ncol = 2,
  byrow = TRUE
  )
)
```

## Clustering

```{r}
seurat_obj <- FindNeighbors(
  seurat_obj, 
  reduction = "integrated.cca",
)
seurat_obj <- FindClusters(
  seurat_obj,
  resolution = .4,
  cluster.name = "cca_clusters"
)
```

## Visualization

```{r}
seeds <- c(42, 17, 37, 97, 31)
for (seed in seeds) {
  seurat_obj <- RunUMAP(
    seurat_obj,
    reduction = "integrated.cca",
    dims = 1:30,
    reduction.name = sprintf("umap.cca_seed%s", seed),
    metric = "euclidean",
    n.neighbors = 30,
    min.dist = .1,
    seed.use = seed
  )
}
```

Show UMAP embedding of different seeds

```{r}
for (seed in seeds) {
  p <- DimPlot(
    seurat_obj,
    reduction = sprintf("umap.cca_seed%s", seed)
  )
  print(p)
}
```


```{r}
p_treatment <- DimPlot(
  seurat_obj,
  reduction = sprintf(
    "umap.cca_seed%s", seeds[2]
  ),
  group.by = "treatment",
  combine = FALSE,
  label.size = 2
)
p_individual <- DimPlot(
  seurat_obj,
  reduction = sprintf(
    "umap.cca_seed%s", seeds[2]
  ),
  group.by = "individual",
  combine = FALSE,
  label.size = 2
)
p_rpca_clusters <- DimPlot(
  seurat_obj,
  reduction = sprintf(
    "umap.cca_seed%s", seeds[2]
  ),
  group.by = "cca_clusters",
  combine = FALSE,
  label.size = 2
)
print(p_treatment[[1]])
print(p_individual[[1]])
print(p_rpca_clusters[[1]])
```

## Show composition of treatment sample per cluster

```{r}
plot_data <- seurat_obj@meta.data %>%
  select(treatment, cca_clusters) %>%
  mutate(cca_clusters = factor(cca_clusters))

ggplot(plot_data, aes(x = cca_clusters, fill = treatment)) +
  geom_bar(position = "fill") +
  ylab("Proportion") +
  xlab("Cluster") +
  theme_classic()
```

## Plot marker gene expression

HD-ZIP III (Group #340, Tung et al., 2023)

```{r}
FeaturePlot(
  seurat_obj,
  reduction = sprintf(
    "umap.cca_seed%s", seeds[2]
  ),
  features = c(
    "Potri.018G045100.v4.1"
  )
)
```

Fusiform Organizer

```{r}
FeaturePlot(
  seurat_obj,
  reduction = sprintf(
    "umap.cca_seed%s", seeds[2]
  ),
  features = c(
    "Potri.001G154100.v4.1", 
    "Potri.015G048700.v4.1"
  ),
  combine = FALSE
)
```
