---
params:
  path_cca_embedding: str
  path_cell_type_csv: str
  path_pseudo_time: str
  path_lineage_weight: str
  path_outdir: str
---

NOTE: Run with docker image chiaenu/rmd-slingshot:2.4.0

This R markdown is used to perform pseudo-time inference using Slingshot 2.4.0 on
Seurat integrated CCA embeddings following this tutorial: 
https://nbisweden.github.io/workshop-archive/workshop-scRNAseq/2020-01-27/labs/compiled/slingshot/slingshot.html

# Setups

```{r Setups}
library(slingshot)
library(magrittr)

cell_order <- c("FuO", "RO", "FuEP", "RP", "FuIP", "Ray", "Vessel", "Fiber")
palette <- c(
  "#9467BD",
  "#FF7F0F",
  "#2AA02A",
  "#F8E71C",
  "#8C564B",
  "#E377C2",
  "#1F77B4",
  "#D62728",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4",
  "#d4d4d4"
)
```

## Parse Rmd parameters

```{r Parse Rmd parameters}
# debug only:
# path_cca_embedding <- "outputs/notebooks-new/Seurat_integration/cca_embedding.csv"
# path_cell_type_csv <- "outputs/notebooks-new/Cell_type_annotation/cell_type_annotation.csv"
# path_pseudo_time <- "outputs/notebooks-new/Slingshot/pseudotime.csv"
# path_lineage_weight <- "outputs/notebooks-new/Slingshot/weight.csv"
# path_outdir <- "outputs/notebooks-new/Slingshot"

path_cca_embedding <- params$path_cca_embedding
path_cell_type_csv <- params$path_cell_type_csv
path_pseudo_time <- params$path_pseudo_time
path_lineage_weight <- params$path_lineage_weight
path_outdir <- params$path_outdir
```

Create output directory

```{r Create output directory}
dir.create(path_outdir, recursive = TRUE, showWarnings = FALSE)
```

# Analysis

Read data

```{r Read data}
# clusters
foo <- read.csv(path_cell_type_csv, row.names = 1)
clusters <- foo$cell_label
clusters[!clusters %in% cell_order] <- "-1" # mark unknown cell types as -1
names(clusters) <- row.names(foo)
clusters <- as.factor(clusters)

# embedding
emb <- read.csv(path_cca_embedding, row.names = 1)
```

## Run Slingshot

Construct lineage object

```{r Construct lineage}
lineages <- new("SlingshotDataSet")

# add cell embedding
lineages@reducedDim <- as.matrix(emb)

# add cluster matrix
mtx_clus <- matrix(0, nrow=length(clusters), ncol=length(levels(clusters)))
row.names(mtx_clus) <- names(clusters)
colnames(mtx_clus) <- levels(clusters)
for (i in seq(length(clusters))) {
  mtx_clus[names(clusters)[i], clusters[i]] <- 1
}
lineages@clusterLabels <- mtx_clus

# add lineages
lineages@lineages <- list(
  "Lineage1" = c("FuO", "FuEP", "FuIP", "Fiber"),
  "Lineage2" = c("RO", "RP", "Ray")
)

# add adjacency matrix
mtx_adj <- matrix(0, nrow = length(levels(clusters)), ncol = length(levels(clusters)))
row.names(mtx_adj) <- colnames(mtx_adj) <- levels(clusters)
for (l_i in lineages@lineages) {
  for (c_i in seq(length(l_i) - 1)) {
    mtx_adj[l_i[c_i], l_i[c_i + 1]] <- 1
    mtx_adj[l_i[c_i + 1], l_i[c_i]] <- 1
  }
}
lineages@adjacency <- mtx_adj

# print lineages
print(lineages)
```

Fit curves

```{r getCurves}
curves <- getCurves(
  lineages,
  approx_points = 200
) %>% SlingshotDataSet
```

Pseudo-time inference

```{r Pseudo-time inference}
pseudotime <- slingPseudotime(
    curves,
    na = FALSE
)
weight <- slingCurveWeights(
    curves
)
```

## Export the pseudo-time & lineage-weight matrix 

```{r Export}
write.csv(
  pseudotime,
  path_pseudo_time
)
write.csv(
  weight,
  path_lineage_weight
)
```