---
title: "Identify differentially edited or ASE loci"
author: "Chia-Chen Chu"
date: '2025-05-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
```{r}
library(Matrix)
library(data.table)
library(ggplot2)
library(magrittr)
library(doParallel)
library(foreach)
setDTthreads(threads = 60)
```


## Configure work place
```{r}
# output directory
path_output <- "outputs/notebooks/Differential-analysis/"
dir.create(path_output)

# single-cell SNV data
path_alt_mtx <- "outputs/VariantCalling/vartrix/ptr_tenx_batch1/alt.mtx"
path_ref_mtx <- "outputs/VariantCalling/vartrix/ptr_tenx_batch1/ref.mtx"
path_barcodes <- "outputs/Remapping/renamer/ptr_tenx_batch1/barcodes.tsv"
path_snv_loci <- "outputs/VariantCalling/vawk/ptr_tenx_batch1.snv.loci.txt"
# read data
alt_mtx <- readMM(path_alt_mtx)
ref_mtx <- readMM(path_ref_mtx)
barcodes <- fread(path_barcodes, header = FALSE)
snv_loci <- fread(path_snv_loci, header = FALSE)
colnames(alt_mtx) <- barcodes$V1
row.names(alt_mtx) <- snv_loci$V1
colnames(ref_mtx) <- barcodes$V1
row.names(ref_mtx) <- snv_loci$V1
# calculate coverage/alt_frac matrix
cov_thresh <- 5
cov_mtx <- alt_mtx + ref_mtx
alt_frac <- alt_mtx / cov_mtx

# single-cell SNV clusters
path_snv_cluster <- "outputs/notebooks/UMAP-show-scSNV/snv-cluster.csv"
snv_cluster <- fread(path_snv_cluster, header = TRUE, sep = ",")
all(snv_cluster$Barcode == colnames(alt_mtx)) # TRUE

# gene-expression clusters
path_cell_types <- "references/ptr_tenx_batch1_rs17_curated.csv"
ge_cluster_4860 <- fread(path_cell_types, header = TRUE, sep = ",")
setkey(ge_cluster_4860, Barcode)
ge_cluster_4870 <- ge_cluster_4860[colnames(alt_mtx), ]

# R|R loci
path_hom_ref <- "outputs/VariantCalling-DNA/gatk_joint/hom_ref.vcf"
hom_ref <- fread(cmd=sprintf("grep -v '^##' %s", path_hom_ref))
hom_ref[, POS := trimws(POS)]
hom_ref[, RN := paste0(`#CHROM`, ":", POS)]


# REDITs scripts
path_REDIT_LLR <- "src/REDITs/REDIT_LLR.R"
source(path_REDIT_LLR)

# vars
perc_thresh <- 0.01
```

## Show SNV cluster colors
```{r}
color_snv <- c(
    "1" = "#197D95",
    "2" = "#4DBF92",
    "3" = "#6362A7",
    "4" = "#06185B",
    "5" = "#72E993",
    "6" = "#DEF5E5",
    "7" = "#6A0B8C",
    "8" = "#3C91E6",
    "9" = "#F4A261",
    "10" = "#E76F51"
)

# Base R color bar
barplot(
    rep(1, length(color_snv)),
    col = color_snv,
    names.arg = names(color_snv),
    las = 2,
    cex.names = 0.8,
    border = NA,
    main = "Color Palette for SNV Clusters"
)
```
## Show GE cluster colors
```{r}
color_ge <- ge_cluster[, c("Cluster", "Color")] %>% unique()
color_ge <- color_ge[order(color_ge$Cluster), ]
# Base R color bar
barplot(
    rep(1, nrow(color_ge)),
    col = color_ge$Color,
    names.arg = color_ge$Cluster,
    las = 2,
    cex.names = 0.8,
    border = NA,
    main = "Color Palette for GE Clusters"
)
```
## Plot distribution of coverage
```{r}
svg(sprintf("%s/hist_coverage_locus=all.svg", path_output))
hist(
    as.vector(cov_mtx[cov_mtx > 0]),
    xlab = "coverage",
    ylab = "No. locus-cell",
    main = "Coverage at each locus-cell",
    xlim = c(0, 20),
    breaks = c(1:500)
)
dev.off()
```
```{r}
svg(sprintf("%s/hist_coverage_locus_cov_gt_10.svg", path_output))
hist(
    as.vector(cov_mtx[cov_mtx > 10]),
    xlab = "coverage",
    ylab = "No. locus-cell",
    main = "Coverage at each locus-cell (coverage > 10)",
    xlim = c(0, 100),
    breaks = c(1:500)
)
dev.off()
```

## N_cells(coverage >= cov_thresh) at each locus
```{r}
n_cell_cov_gt_thresh <- rowSums(cov_mtx >= cov_thresh)
svg(sprintf("%s/hist_n_cells_coverage_gt_%s.svg", path_output, cov_thresh))
hist(
    n_cell_cov_gt_thresh[n_cell_cov_gt_thresh > 0],
    xlab = "N_cells(coverage > threshold)",
    ylab = "N_locus",
    main = "N_cells(coverage > threshold) at each locus",
    breaks = 0:5000,
    xlim = c(0, 50)
)
dev.off()
```



## Test for differential editing using REDITs
```{r}
cond1_clusters <- c(7)
cond2_clusters <- c(1)
cond1_label <- "Fiber"
cond2_label <- "Vessel"
fill_manual <- c("Fiber" = "red", "Vessel" = "blue")

# this data.table store differential analysis results
da_results <- data.table(
    locus = hom_ref$RN,
    perc_cond1 = NA_real_,
    perc_cond2 = NA_real_,
    p_value = NA_real_
)
setkey(da_results, locus)

# set group label by cluster number
ge_cluster_4870_copy <- ge_cluster_4870
ge_cluster_4870_copy$groups <- ""
ge_cluster_4870_copy[Cluster %in% cond1_clusters, "groups"] <- "cond1"
ge_cluster_4870_copy[Cluster %in% cond2_clusters, "groups"] <- "cond2"
groups <- ge_cluster_4870_copy$groups
groups_factor <- factor(groups, levels = c("cond1", "cond2"), labels = c(cond1_label, cond2_label))

# subset matrices
alt_mtx_in_group <- alt_mtx[, !is.na(groups_factor)]
ref_mtx_in_group <- ref_mtx[, !is.na(groups_factor)]
cov_mtx_in_group <- cov_mtx[, !is.na(groups_factor)]
alt_frac_in_group <- alt_frac[, !is.na(groups_factor)]

# At each locus calculate Percent(coverage >= cov_thresh) of cond 1 and cond 2
da_results$perc_cond1 <- (cov_mtx[da_results$locus, groups == "cond1"] >= cov_thresh) %>% rowMeans
da_results$perc_cond2 <- (cov_mtx[da_results$locus, groups == "cond2"] >= cov_thresh) %>% rowMeans

# calculate p-value
da_results[(perc_cond1 > perc_thresh) & (perc_cond2 > perc_thresh), p_value := {
    data_valid <- cov_mtx_in_group[.BY$locus, ] >= cov_thresh
    REDIT_LLR(
        data = rbind(
            alt_mtx_in_group[.BY$locus, data_valid],
            ref_mtx_in_group[.BY$locus, data_valid]
        ),
        groups = na.omit(groups_factor)[data_valid] %>% as.vector
    )$p.value
}, by = locus]

# correct p-value
da_results[!is.na(p_value), p_adj := p.adjust(p_value, method = "BH")]

# extract differential analysis results
to_plot <- da_results[!is.na(p_adj), ]

# plot alt-fraction per locus
for (nth_row in 1:nrow(to_plot)) {
    # locus name
    locus <- to_plot[nth_row, locus]
    
    # get alt-fraction & coverage
    alt_frac_locus <- data.table(
        alt_frac = alt_frac_in_group[locus, ],
        cov = cov_mtx_in_group[locus, ],
        group = na.omit(groups_factor)
    )
    
    # plot only coverage > threshold
    alt_frac_locus <- alt_frac_locus[cov >= cov_thresh, ]
    p <- ggplot(alt_frac_locus, aes(x = alt_frac, fill = group)) + 
        scale_fill_manual(
            values = fill_manual
        ) +
        geom_histogram(
            aes(y = after_stat(density)),
            bins = 100,
            position = "identity",
            alpha = 0.5
        ) +
        xlim(-0.01, 1.01) +
        labs(
            x = "Alternative Allele Fraction",
            y = "Density",
            title = "Normalized Histogram of alt_frac at %s (R|R)\np-adj=%.3f\n%%%s=%.1f%%\n%%%s=%.1f%%\ncoverage>=%s" %>% sprintf(
                locus,
                to_plot[nth_row, p_adj],
                cond1_label,
                to_plot[nth_row, perc_cond1] * 100,
                cond2_label,
                to_plot[nth_row, perc_cond2] * 100,
                cov_thresh
            ),
            fill = "Group"
        ) +
        theme_minimal()
    svg(sprintf(
        "%s/hist_alt_frac_%s_vs_%s_%s.svg",
        path_output,
        cond1_label,
        cond2_label,
        locus %>% gsub(":", "-", .)
    ))
    print(p)
    dev.off()
}
```


Test differential editing between ray cells and all other cells using REDITs
```{r}
# this data.table store differential analysis results
da_results_ray_to_all <- data.table(
    locus = hom_ref$RN,
    perc_cond1 = NA_real_,
    perc_cond2 = NA_real_,
    pval = NA_real_
)
setkey(da_results_ray_to_all, locus)

# set group label by cluster number
cond1 <- c(9, 4)
cond2 <- setdiff(1:10, cond1)
snv_cluster_copy <- snv_cluster
snv_cluster_copy$groups <- ""
snv_cluster_copy[Cluster %in% cond1, "groups"] <- "cond1"
snv_cluster_copy[Cluster %in% cond2, "groups"] <- "cond2"
groups <- snv_cluster_copy$groups
groups_factor <- factor(groups)

# calculate percent non-NA occurrence of cond 1 and cond 2
da_results_ray_to_all$perc_cond1 <- (cov_mtx >= cov_thresh)[da_results_ray_to_all$locus, groups == "cond1"] %>% rowMeans
da_results_ray_to_all$perc_cond2 <- (cov_mtx >= cov_thresh)[da_results_ray_to_all$locus, groups == "cond2"] %>% rowMeans

# calculate p-value 
alt_mtx_in_group <- alt_mtx[, snv_cluster[Cluster %in% c(cond1, cond2), Barcode]]
ref_mtx_in_group <- ref_mtx[, snv_cluster[Cluster %in% c(cond1, cond2), Barcode]]
da_results_ray_to_all[(perc_cond1 > perc_thresh) & (perc_cond2 > perc_thresh), p_value := {
    data <- rbind(alt_mtx_in_group[.BY$locus, ], ref_mtx_in_group[.BY$locus, ])
    groups <- snv_cluster_copy[Cluster %in% c(cond1, cond2), groups] %>% as.character
    data_valid <- colSums(data) >= cov_thresh
    data <- data[, data_valid]
    groups <- groups[data_valid]
    REDIT_LLR(data ,groups)$p.value
}, by = locus]


# correct p-value
da_results_ray_to_all[, p_adj := p.adjust(p_value, method = "BH")]
```


Plot alt_frac marked with p-values
```{r}
to_plot <- da_results_ray_to_all[!is.na(p_adj),]
alt_frac_in_group <- alt_frac[, snv_cluster[Cluster %in% c(cond1, cond2), Barcode]]
cov_mtx_in_group <- cov_mtx[, snv_cluster[Cluster %in% c(cond1, cond2), Barcode]]
for (nth_row in 1:nrow(to_plot)) {
    locus <- to_plot[nth_row, locus]
    # get alt-fraction
    alt_frac_locus <- data.table(
        alt_frac = alt_frac_in_group[locus, ],
        cov = cov_mtx_in_group[locus, ],
        group = snv_cluster_copy[Cluster %in% c(cond1, cond2), groups] %>% as.character
    )
    # plot only coverage > threshold
    alt_frac_locus <- alt_frac_locus[cov >= cov_thresh, ]
    p <- ggplot(alt_frac_locus, 
            aes(
                x = alt_frac,
                fill = factor(group, levels = c("cond1", "cond2"), labels = c("RO", "others"))
            )
        ) + 
        scale_fill_manual(
            values = c(
                "RO" = "red",
                "others" = "blue"
                )
        ) +
        geom_histogram(
            aes(y = after_stat(density)),
            bins = 100,
            position = "identity",
            alpha = 0.5
        ) +
        xlim(-0.01, 1.01) +
        labs(
            x = "Alternative Allele Fraction",
            y = "Density",
            title = "Normalized Histogram of alt_frac at %s (R|R)\np-adj=%s\nperc_cond1=%s\nperc_cond2=%s\ncov_thresh>=%s" %>% sprintf(
                locus,
                to_plot[nth_row, p_adj],
                to_plot[nth_row, perc_cond1],
                to_plot[nth_row, perc_cond2],
                cov_thresh
            ),
            fill = "Group"
        ) +
        theme_minimal()
    svg(sprintf(
        "%s/hist_alt_frac_RO:Others_%s.svg",
        path_output,
        locus
    ))
    print(p)
    dev.off()
}
```


## Re-write differential analysis (& plotting?) into 1 function
```{r}
```



