```{r}
library(Seurat)
library(Matrix)
```

Functions
```{r}
read10x_to_seurat <- function(path_10x, sample_name) {
  CreateSeuratObject(counts = Read10X(path_10x), project = sample_name)
}
```

Define path, vars
```{r}
# vars
SAMPLES <- c(
  "ptr_tenx_batch1",
  "ptr_tenx_batch2",
  "ptr_tenx_tsv1",
  "ptr_tenx_tsv2",
  "ptr_tenx_tsv3",
  "ptr_tenx_tsv4",
  "ptr_tenx_tsv5",
  "ptr_tenx_tst1",
  "ptr_tenx_tst2",
  "ptr_tenx_tst3",
  "ptr_tenx_tst4",
  "ptr_tenx_tso1",
  "ptr_tenx_tso2",
  "ptr_tenx_tso3",
  "ptr_tenx_tso4"
)
cluster_colors <- c(
    "1"="#1F77B4",
    "2"="#7A4A33",
    "3"="#E28413",
    "4"="#F3EC5B",
    "5"="#61C9A8",
    "6"="#6362A7",
    "7"="#FF0000",
    "8"="#FF8F8F",
    "9"="#D4D4D4",
    "10"="#D4D4D4",
    "11"="#D4D4D4",
    "12"="#D4D4D4"
)

# params
NORMALIZATION.METHOD <- "LogNormalize"
SCALE.FATCOR <- 10000
REFERENCE <- 1 # sample 1 --> ptr_tenx_batch1
METHOD <- CCAIntegration
NEW.REDUCTION <- "integrated.cca"

# paths
path_output <- "outputs/notebooks/Seurat-integrate-Ptr-samples/20250729_172935"
path_cell_identity <- "references/ptr_tenx_batch1_rs17_curated.csv"
path_cell_identity_by_NN <- "outputs/notebooks/Seurat-integrate-Ptr-samples/cell_identity_all.csv"
path_cellranger_mtx <- function(sample) {sprintf("outputs/CellRanger/count/%s/outs/filtered_feature_bc_matrix/", sample)}
path_seurat_obj <- "outputs/notebooks/Seurat-integrate-Ptr-samples/seurat_obj.rds"
path_integrated_embedding <- "outputs/notebooks/Seurat-integrate-Ptr-samples/integrated.cca_ref=1_LogNorm_scaleFactor=10000.csv"
path_integrated_count <- "outputs/notebooks/Seurat-integrate-Ptr-samples/integrated_count.cca_ref=1_LogNorm_scaleFactor=10000.mtx"
path_integrated_data <- "outputs/notebooks/Seurat-integrate-Ptr-samples/integrated_data.cca_ref=1_LogNorm_scaleFactor=10000.mtx"
path_integrated_scaled_data <- "outputs/notebooks/Seurat-integrate-Ptr-samples/integrated_scaled_data.cca_ref=1_LogNorm_scaleFactor=10000.mtx"
path_median_umap <- "outputs/notebooks/Seurat-integrate-Ptr-samples/20250729_172935/median_umap.csv"
path_dre_results <- "outputs/notebooks/Differential-analysis-multi/20250729_172935/DRE_complete_table.csv"

# load/create paths
dir.create(path_output)
# cell_identity_by_NN <- read.csv(path_cell_identity_by_NN)
seurat_objs <- mapply(read10x_to_seurat, path_cellranger_mtx(SAMPLES), SAMPLES)
seurat_obj <- merge(x = seurat_objs[[1]], y = seurat_objs[2:length(seurat_objs)])
rm(seurat_objs); gc()
```

### Integrate
Normalize data
```{r}
seurat_obj <- NormalizeData(
  seurat_obj, 
  normalization.method = NORMALIZATION.METHOD, 
  scale.factor = SCALE.FATCOR
)
gc()
# saveRDS(seurat_obj, file = path_seurat_obj) # save save
```

Run PCA
```{r}
seurat_obj <- FindVariableFeatures(
  seurat_obj,
  selection.method = "vst",
  nfeatures = 2000
)
seurat_obj <- ScaleData(
  seurat_obj,
  do.scale = TRUE,
  do.center = TRUE
)
seurat_obj <- RunPCA(seurat_obj)
gc()
# saveRDS(seurat_obj, file = path_seurat_obj) # save save
```

### Integrate based on normalized expression using CCA
Layer `ptr_tenx_batch1` was used as reference
This will correct the original reduction: PCA
```{r}
# run integration
options(future.globals.maxSize = 10000 * 1024^2)
seurat_obj <- IntegrateLayers(
  object = seurat_obj,
  method = METHOD, 
  orig.reduction = "pca", 
  new.reduction = NEW.REDUCTION,
  reference = REFERENCE,
  verbose = TRUE
)
seurat_obj[["RNA"]] <- JoinLayers(seurat_obj[["RNA"]])
gc()
```

### Write Seurat object/PCA reduction to file
```{r}
# saveRDS(seurat_obj, file = path_seurat_obj)
write.csv(
  x = Embeddings(seurat_obj, reduction = "integrated.cca"),
  file = path_integrated_embedding,
  quote = FALSE,
  row.names = TRUE
)
sparse_matrix_data <- GetAssayData(object = seurat_obj, assay = "RNA", layer = "data")
writeMM(obj = sparse_matrix_data, file = path_integrated_data)
```

### Find cluster
```{r}
# seurat_obj <- FindNeighbors(seurat_obj, reduction = NEW.REDUCTION, dims = 1:30)
# seurat_obj <- FindClusters(seurat_obj, resolution = 1)
```

### Import median UMAP embedding
```{r}
median_umap <- read.csv(path_median_umap)
umap_coords_matrix <- as.matrix(median_umap[,c("x1", "x2")])
row.names(umap_coords_matrix) <- median_umap$Barcode
umap_dr <- CreateDimReducObject(embeddings = umap_coords_matrix, key = "UMAP_", global = TRUE, assay = "RNA")

# Add the DimReducObject to your Seurat object
seurat_obj[["umap"]] <- umap_dr
```

### Plotting
```{r}
seurat_obj$orig.ident <- factor(seurat_obj$orig.ident, level = SAMPLES)
```


```{r}
hd_zip3 <- c(
  "Potri.001G188800.v4.1",
  "Potri.001G372300.v4.1",
  "Potri.003G050100.v4.1",
  "Potri.004G211300.v4.1",
  "Potri.006G237500.v4.1",
  "Potri.009G014500.v4.1",
  "Potri.011G098300.v4.1",
  "Potri.018G045100.v4.1"
)
for (g in hd_zip3) {
  png(
    sprintf("%s/gene_expression/hd_zip3_%s.png", path_output, g),
    height = 2000,
    width = 2000 * 15,
    res = 200
  )
  p <- FeaturePlot(
    seurat_obj,
    features = g,
    reduction = "umap",
    split.by = "orig.ident",
    ncol = 5,
    pt.size = 1,
    slot = "data"
  )
  print(p)
  dev.off()
}
```


### Show DRE on UMAP (gene expression)
```{r}
dre_results <- read.csv(path_dre_results)
DRE_genes <- dre_results[order(dre_results$p_adj)[1:70], ]$GENEID
DRE_genes <- DRE_genes[nchar(DRE_genes) > 0]
DRE_genes <- paste0(DRE_genes, ".v4.1")
DRE_genes <- DRE_genes[order(DRE_genes)]
for (geneid in DRE_genes) {
  p <- FeaturePlot(
    seurat_obj, 
    features = geneid,
    split.by="orig.ident",
    order = TRUE,
    pt.size = 0.5,
    label = FALSE,
  ) + RestoreLegend()
  png(
    sprintf("%s/gene_expression/%s.png", path_output, geneid),
    height = 2000,
    width = 2000 * 15,
    res = 200
  )
  print(p)
  dev.off()
}

```

```{r}
summary(FetchData(seurat_obj, vars = "Potri.001G268400.v4.1", layer = "count"))
```











### Plot UMAP colored by Seurat cluster (split by batch)
```{r}
png(
  sprintf("%s/%s", path_output, "umap_log_cca_ref=1_split=batch_col=seurat_cluster.png"),
  height = 1600 * 3,
  width = 1600 * 5,
  res = 300
)
DimPlot(
  seurat_obj,
  reduction = "umap",
  split.by = "orig.ident",
  pt.size = 2,
  ncol = 5,
)
dev.off()
```


### Write Seurat object/PCA reduction to file
```{r}
# saveRDS(seurat_obj, file = path_seurat_obj)
write.csv(
  x = Embeddings(seurat_obj, reduction = "integrated.cca"),
  file = path_integrated_embedding,
  quote = FALSE,
  row.names = TRUE
)
```

### Plot UMAP colored by NN-transfered cell type
```{r}
seurat_obj$cell_type <- cell_identity_by_NN$Cluster
Idents(seurat_obj) <- "cell_type"
png(
  sprintf("%s/%s", path_output, "umap_rpca_ref=1_split=batch_col=reference_nn.png"),
  height = 1600 * 3,
  width = 1600 * 5,
  res = 300
)
DimPlot(
  seurat_obj,
  reduction = "umap",
  split.by = "orig.ident",
  cols = cluster_colors,
  pt.size = 2,
  ncol = 5,
)
dev.off()
```

