```{r}
library(Seurat)
library(magrittr)
library(data.table)
library(Matrix)
```

Define data paths
```{r}
sample_names <- c(
  "ptr_tenx_batch1",
  "ptr_tenx_batch2",
  "ptr_tenx_tsv1",
  "ptr_tenx_tsv2",
  "ptr_tenx_tsv3",
  "ptr_tenx_tsv4",
  "ptr_tenx_tsv5"
)
cluster_colors <- c(
    "1"="#1F77B4",
    "2"="#7A4A33",
    "3"="#E28413",
    "4"="#F3EC5B",
    "5"="#61C9A8",
    "6"="#6362A7",
    "7"="#FF0000",
    "8"="#FF8F8F",
    "9"="#D4D4D4",
    "10"="#D4D4D4",
    "11"="#D4D4D4",
    "12"="#D4D4D4"
)

# data paths
path_output <- "outputs/notebooks/Show-DRE-sites-on-Seurat-UMAP/"
path_cell_identity <- "references/ptr_tenx_batch1_rs17_curated.csv"
path_DRE_RO_FuO <- "outputs/notebooks/Differential-analysis/DRES_RO_vs_FuO.csv"
path_DRE_RO_Others <- "outputs/notebooks/Differential-analysis/DRES_RO_vs_Others.csv"
path_editing_level <- "outputs/notebooks/Show-DRE-sites-on-Seurat-UMAP/editing_level.csv"
path_variant_annotation <- "tests/test-VariantAnnotation/variant_annotation_hom_ref.csv"

# Read cell identity
cell_identity <- read.csv(path_cell_identity)
rownames(cell_identity) <- cell_identity$Barcode
cell_identity$Cluster_name <- cluster_names[cell_identity$Cluster]
head(cell_identity)

# Read DRE results
DRE_RO_FuO <- read.csv(path_DRE_RO_FuO)
DRE_RO_Others <- read.csv(path_DRE_RO_Others)
cat("#RO_vs_FuO: ", nrow(DRE_RO_FuO), "\n")
cat("#RO_vs_others: ", nrow(DRE_RO_Others), "\n")
cat("Union: ", union(DRE_RO_FuO$locus, DRE_RO_Others$locus) %>% length, "\n")
cat("Intersect: ", intersect(DRE_RO_FuO$locus, DRE_RO_Others$locus) %>% length, "\n")
DRES_RO_FuO_Others <- union(DRE_RO_FuO$locus, DRE_RO_Others$locus)

# Read editing levels at DRE sites (calculated using Python)
editing_level <- read.csv(path_editing_level,row.names = 1)

# Read variant annotation
variant_annotation <- read.csv(path_variant_annotation)

# Read scSNV data
get_alt_frac_mtx <- function(sample_name, loci_of_interest, cov_threshold) {
  # data path based on sample name
  path_barcodes <- "outputs/Remapping/renamer/%s/barcodes.tsv" %>% sprintf(., sample_name)
  path_snv_loci <- "outputs/VariantCalling/vawk/%s.snv.loci.txt" %>% sprintf(., sample_name)
  path_alt_mtx <- "outputs/VariantCalling/vartrix/%s/alt.mtx" %>% sprintf(., sample_name)
  path_ref_mtx <- "outputs/VariantCalling/vartrix/%s/ref.mtx" %>% sprintf(., sample_name)
  
  # read matrix
  barcodes <- path_barcodes %>% read.table(header = FALSE)
  snv_loci <- path_snv_loci %>% read.table(header = FALSE)
  alt_mtx <- path_alt_mtx %>% readMM # %>% as.matrix %>% as.data.frame
  ref_mtx <- path_ref_mtx %>% readMM # %>% as.matrix %>% as.data.frame
  
  # set dim names
  colnames(alt_mtx) <- barcodes$V1
  colnames(ref_mtx) <- barcodes$V1
  row.names(alt_mtx) <- snv_loci$V1
  row.names(ref_mtx) <- snv_loci$V1
  
  # subset to loci of interest
  alt_mtx <- alt_mtx[intersect(row.names(alt_mtx), validated_sites), ]
  ref_mtx <- ref_mtx[intersect(row.names(ref_mtx), validated_sites), ]
  cov_mtx <- alt_mtx + ref_mtx
  alt_frac <- alt_mtx / cov_mtx
  
  # fill NA with 0 and mark where coverage < 5 as 0
  alt_frac[is.na(alt_frac)] <- 0
  alt_frac[cov_mtx < 5] <- 0
  
  # return 
  return(alt_frac)
}
alt_fracs <- sapply(sample_names, get_alt_frac_mtx, validated_sites, 5)
```

Read 10x data
```{r}
read10x_to_seurat <- function(sample_name) {
  path_cellranger_mtx <- sprintf(
    "outputs/CellRanger/count/%s/outs/filtered_feature_bc_matrix/", 
    sample_name
  )
  return(CreateSeuratObject(counts = Read10X(path_cellranger_mtx), project = sample_name))
}
seurat_objs <- sapply(sample_names, read10x_to_seurat)
seurat_obj <- merge(x = seurat_objs[[1]], y = seurat_objs[2:length(seurat_objs)])
# rm(seurat_objs)
```

Integration
```{r}
seurat_obj <- SCTransform(seurat_obj)
seurat_obj <- RunPCA(seurat_obj)
seurat_obj <- IntegrateLayers(
  object = seurat_obj, 
  method = CCAIntegration, 
  orig.reduction = "pca", 
  new.reduction = "integrated.cca",
  normalization.method = "SCT",
  verbose = FALSE
)
seurat_obj[["RNA"]] <- JoinLayers(seurat_obj[["RNA"]])
```

Find clusters
```{r}
seurat_obj <- FindNeighbors(seurat_obj, reduction = "integrated.cca", dims = 1:30)
seurat_obj <- FindClusters(seurat_obj, resolution = 1)
```

Give cell identity by the nearest ptx_tenx_batch1 cell on the integrated reductions
```{r}
reductions <- Embeddings(seurat_obj, reduction = "integrated.cca")[,1:30]
write.csv(
  reductions,
  sprintf("%s/%s", path_output, "integrated.cca.csv"),
)
```
Run a python script...
```{r}
cell_identity <- read.csv(sprintf("%s/%s", path_output, "cell_identity_all.csv"))
cell_identity$Cluster <- as.character(cell_identity$Cluster)
rownames(cell_identity) <- cell_identity$Barcode
```

UMAP
```{r}
seurat_obj <- RunUMAP(
  seurat_obj,
  dims = 1:30, 
  reduction = "integrated.cca",
  seed.use = 42,
  n.neighbors = 10L,
  n.components = 2L,
  metric = "correlation",
  min.dist = 0.15,
)
```

Plot UMAP with NN-inferred cell type
```{r}
seurat_obj$celltype <- cell_identity$Cluster
Idents(seurat_obj) <- "celltype"
# Visualization
png(sprintf("%s/%s", path_output, "umap.png"), height = 1600, width = 1600, res = 300)
DimPlot(
  seurat_obj,
  reduction = "umap",
  cols = cluster_colors,
) + NoAxes() + NoLegend()
dev.off()
```

Plot UMAP with NN-inferred cell type, splited by batch
```{r}
seurat_obj$celltype <- cell_identity$Cluster
Idents(seurat_obj) <- "celltype"
# Visualization
png(sprintf("%s/%s", path_output, "umap_splited.png"), height = 1600, width = 10000, res = 300)
DimPlot(
  seurat_obj,
  reduction = "umap",
  cols = cluster_colors,
  split.by="orig.ident"
) + NoAxes() + NoLegend()
dev.off()
```


Plot expression of genes with missense mutation validated by proteome
```{r}
DRE_genes <- editing_level %>% 
  colnames %>% 
  grep("_", ., value = TRUE) %>% 
  sapply(., function(x){
    unlist(strsplit(x, "_"))[2] 
  })
for (geneid in DRE_genes) {
  p <- FeaturePlot(
    seurat_obj, 
    features = geneid,
    split.by="orig.ident",
    order = TRUE,
    pt.size = 0.5,
    label = FALSE,
  ) + RestoreLegend()
  png(
    sprintf("%s/gene_expression/%s.png", path_output, geneid),
    height = 1600,
    width = 12000,
    res = 300
  )
  print(p)
  dev.off()
}
```

# Plot editing level of the missense sites
```{r}
seurat_obj <- AddMetaData(seurat_obj, editing_level)
```

```{r}
for (locus in colnames(editing_level)) {
  png(
    sprintf(
      "%s/editing_level/%s.png",
      path_output, 
      locus
    ), 
    height = 1600,
    width = 12000,
    res = 300
  )
  p <- FeaturePlot(
    seurat_obj,
    reduction = "umap",
    features = locus,
    order = TRUE,
    split.by = "orig.ident",
    pt.size = 0.5,
  ) + RestoreLegend()
  print(p)
  dev.off()
}
```

Editing level * expression level
```{r}
geneids <- colnames(editing_level) %>% sapply(
  ., function(x){unlist(strsplit(x, "_"))[2]}
)
gene_expression <- FetchData(seurat_obj, vars = geneids)

rm("edited_expression")
for (locus_geneid in colnames(editing_level)) {
  expression_level_gene <- gene_expression[unlist(strsplit(locus_geneid, "_"))[2]]
  if (!exists("edited_expression")) {
    edited_expression <- editing_level[locus_geneid] * expression_level_gene
  } else {
    edited_expression <- cbind(
      edited_expression,
      editing_level[locus_geneid] * expression_level_gene
    )  
  }
}
colnames(edited_expression) <- paste0(colnames(edited_expression), "_edited_expression")
seurat_obj <- AddMetaData(seurat_obj, edited_expression)
```

Plot
```{r}
for (locus in colnames(edited_expression)) {
  png(
    sprintf(
      "%s/edited_expression_level/%s.png",
      path_output, 
      locus
    ), 
    height = 1600,
    width = 12000,
    res = 300
  )
  p <- FeaturePlot(
    seurat_obj,
    reduction = "umap",
    features = locus,
    order = TRUE,
    split.by = "orig.ident",
    pt.size = 0.5,
  ) + RestoreLegend()
  print(p)
  dev.off()
}
```


