# run with: docker run --rm -it -v \$(pwd):/data -w /data satijalab/seurat:5.0.0 R

# working directory: snakemake main directory

Load libraries
```{R}
library(Seurat)
library(Matrix)
library(stringr)
library(magrittr)
library(uwot)
library(data.table)
library(dplyr)
```

Create output folder
```{r}
path_output <- "outputs/notebooks/UMAP-show-scSNV-Egr"
dir.create(path_output)
```

Load RNA polymorphism data
```{R}
# paths
path_barcodes <- "outputs/Remapping/renamer/egr_tenx_batch1/barcodes.tsv"
path_snv_loci <- "outputs/VariantCalling/vawk/egr_tenx_batch1.snv.loci.txt"
path_sc_vcf <- "outputs/VariantCalling/bcftools_concat/egr_tenx_batch1.vcf"
path_alt_mtx <- "outputs/VariantCalling/vartrix/egr_tenx_batch1/alt.mtx"
path_ref_mtx <- "outputs/VariantCalling/vartrix/egr_tenx_batch1/ref.mtx"

# load tables
barcodes <- path_barcodes %>% read.table(header = FALSE)
snv_loci <- path_snv_loci %>% read.table(header = FALSE)
sc_vcf <- fread(cmd=sprintf("grep -v '^##' %s", path_sc_vcf)) %>% as.data.frame
sc_vcf$POS %<>% trimws()
rownames(sc_vcf) <- apply(
  sc_vcf[, c("#CHROM", "POS")], 1,
  function(x){paste(x[1], x[2], sep = ":")}
)
# load matrices
alt_mtx <- path_alt_mtx %>% readMM # %>% as.matrix %>% as.data.frame
ref_mtx <- path_ref_mtx %>% readMM # %>% as.matrix %>% as.data.frame
colnames(alt_mtx) <- barcodes$V1
colnames(ref_mtx) <- barcodes$V1
row.names(alt_mtx) <- snv_loci$V1
row.names(ref_mtx) <- snv_loci$V1
```
```{r}
# Calculate alternate fractions (this takes long)
alt_frac <- alt_mtx / (alt_mtx + ref_mtx)
t_alt_frac <- t(alt_frac) # rownames = cells, colnames = loci
```

Load Genotyping data
```{R}
# path
path_hom_alt <- "outputs/VariantCalling-DNA/gatk_joint/egr/hom_alt.vcf"
path_hom_ref <- "outputs/VariantCalling-DNA/gatk_joint/egr/hom_ref.vcf"
path_het <- "outputs/VariantCalling-DNA/gatk_joint/egr/het.vcf"
path_snp <- "outputs/VariantCalling-DNA/gatk_joint/egr/snp.vcf"

# load
hom_alt <- fread(cmd=sprintf("grep -v '^##' %s", path_hom_alt)) %>% as.data.frame
hom_ref <- fread(cmd=sprintf("grep -v '^##' %s", path_hom_ref)) %>% as.data.frame
het <- fread(cmd=sprintf("grep -v '^##' %s", path_het)) %>% as.data.frame
snp <- fread(cmd=sprintf("grep -v '^##' %s", path_snp)) %>% as.data.frame
hom_alt$POS %<>% trimws()
hom_ref$POS %<>% trimws()
het$POS %<>% trimws()
snp$POS %<>% trimws()
rownames(hom_alt) <- apply(
  hom_alt[, c("#CHROM", "POS")], 1,
  function(x){paste(x[1], x[2], sep = ":")}
)
rownames(hom_ref) <- apply(
  hom_ref[, c("#CHROM", "POS")], 1,
  function(x){paste(x[1], x[2], sep = ":")}
)
rownames(het) <- apply(
  het[, c("#CHROM", "POS")], 1,
  function(x){paste(x[1], x[2], sep = ":")}
)
rownames(snp) <- apply(
  snp[, c("#CHROM", "POS")], 1,
  function(x){paste(x[1], x[2], sep = ":")}
)
```

Previous results (data outside of this pipeline)
```{R}
# path
path_cell_types <- "references/seurat_four_trees_cell_identity.csv"

# load data
cell_types <- fread(path_cell_types, header = TRUE)
cell_types <- cell_types[grep("egr_tenx_batch1", Barcode),] # extract only "egr_tenx_batch1" cells
cell_types$Barcode %<>% sub(".*:", "", .) # remove strings prefixing barcode
setkey(cell_types, Barcode) # set column `Barcode` as key
```

Process cell_types and umap_gene_expression data.frame
```{R}
# gene expression cluster colors
color_ge <- c(
    "3"="#E28413",
    "4"="#F3EC5B",
    "8"="#FF8F8F",
    "6"="#6362A7",
    "5"="#61C9A8",
    "2"="#7A4A33",
    "1"="#1F77B4",
    "7"="#FF0000",
    "9"="whitesmoke",
    "10"="whitesmoke",
    "11"="whitesmoke",
    "12"="whitesmoke"
)
# snv cluster colors
color_snv <- c(
    "1" = "#197D95",
    "2" = "#4DBF92",
    "3" = "#6362A7",
    "4" = "#06185B",
    "5" = "#72E993",
    "6" = "#DEF5E5",
    "7" = "#6A0B8C",
    "8" = "#3C91E6",
    "9" = "#F4A261",
    "10" = "#E76F51",
    "11" = "#264653",
    "12" = "#9C89B8",
    "13" = "#F6BD60",
    "14" = "#84A59D",
    "15" = "#F28482"
)

# update cell_types$Color
cell_types$Color <- color_ge[as.character(cell_types$Cluster)] %>% unlist
```

# EDA on alternate fractions
Calculate alternate fraction matrix
```{R}
library(ggplot2)
library(dplyr)
library(tidyr)

# Plot histogram
svg("outputs/notebooks/UMAP-show-scSNV-Egr/alt_frac_hist.svg", width = 7, height = 7)
alt_frac[!is.na(alt_frac)] %>% unlist %>%
    hist(
        main = "Alternate fraction\nAll cells; All loci\nExcluded loci with 0 reads",
        xlab = "Alternate fraction",
        ylab = "# observations at loci-cell",
        breaks = 100
    )
dev.off()
```

Descriptive statistics
```{R}
# calculate variance at each locus while ignoring NAs
locus_var <- apply(t_alt_frac, 2, function(x) {
    if (length(x[!is.na(x)]) > 0) {
        return(var(x, na.rm = TRUE))
    } else {
        return(NA)
    }
})
# calculate mean at each locus while ignoring NAs
locus_mean <- apply(t_alt_frac, 2, function(x) {
    if (length(x[!is.na(x)]) > 0) {
        return(mean(x, na.rm = TRUE))
    } else {
        return(NA)
    }
})
# calculate number of cells at each locus while ignoring NAs
locus_n_cells <- apply(t_alt_frac, 2, function(x) {
    if (length(x[!is.na(x)]) > 0) {
        return(length(x[!is.na(x)]))
    } else {
        return(NA)
    }
})
# plot scatter plot of locus variance v.s. locus n_cells 
# with dot size = locus mean
svg("outputs/notebooks/UMAP-show-scSNV-Egr/locus_variance~n_cells.svg",
    width = 7,
    height = 7,
    bg = "transparent"
)
plot(
    x = locus_n_cells,
    y = locus_var,
    cex = 0.6 * locus_mean,
    main = sprintf(
        "At %s locus expressed by at least 1 cell\nsize=mean(alt_frac)",
        length(locus_var[!is.na(locus_var)])
    ),
    xlab = "Number of cells at locus",
    ylab = "Variance of (alternate fractions) at locus"
)
dev.off()
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/locus_variance~mean.svg",
    width = 7,
    height = 7,
    bg = "transparent"
)
plot(
    x = locus_mean,
    y = locus_var,
    cex = 0.1 * log10(locus_n_cells),
    main = sprintf(
        "At %s locus expressed by at least 1 cell\nsize=log10(n_cells)",
        length(locus_var[!is.na(locus_var)])
    ),
    xlab = "Mean alternate fraction at locus",
    ylab = "Variance of (alternate fractions) at locus"
)
dev.off()
```

Plot histogram of alternate fraction for loci with high variance
```{R}
locus_stats <- data.frame(
    locus_var = locus_var,
    locus_mean = locus_mean,
    locus_n_cells = locus_n_cells
)
# loci with top 10 highest variance & reasonable n_cells
n_cells_cutoff <- 500
for (i in 1:10) {
    locus <- locus_stats %>% 
        subset(locus_n_cells > n_cells_cutoff) %>% 
        arrange(desc(locus_var)) %>%
        slice(i) %>% 
        rownames
    svg(
        sprintf(
            "outputs/notebooks/UMAP-show-scSNV-Egr/alt_frac_hist_%s.svg",
            locus
        ),
        width = 7,
        height = 7,
        bg = "transparent"
    )
    alt_frac_locus <- t_alt_frac[, locus] %>% 
        unlist %>%
        hist(
            main = sprintf(
                "Alternate fraction at locus %s\n%s cells; variance=%s",
                locus,
                sum(!is.na(t_alt_frac[, locus])),
                locus_stats[locus, "locus_var"]
            ),
            xlab = "Alternate fraction",
            ylab = "Observed in N cells",
            breaks = 100
        )
    dev.off()
}
```

Plot histogram of alternate fraction for loci with high cell counts
```{R}
locus_stats <- data.frame(
    locus_var = locus_var,
    locus_mean = locus_mean,
    locus_n_cells = locus_n_cells
)
# loci with top 10 highest n_cells & reasonable variance
for (i in 1:10) {
    locus <- locus_stats %>% 
        arrange(desc(locus_n_cells)) %>%
        slice(i) %>% 
        rownames
    svg(
        sprintf(
            "outputs/notebooks/UMAP-show-scSNV-Egr/alt_frac_hist_%s.svg",
            locus
        ),
        width = 7,
        height = 7,
        bg = "transparent"
    )
    alt_frac_locus <- t_alt_frac[, locus] %>% 
        unlist %>%
        hist(
            main = sprintf(
                "Alternate fraction at locus %s\n%s cells; variance=%s",
                locus,
                sum(!is.na(t_alt_frac[, locus])),
                locus_stats[locus, "locus_var"]
            ),
            xlab = "Alternate fraction",
            ylab = "Observed in N cells",
            breaks = 100
        )
    dev.off()
}
```

# Run UMAP-PCA on alternate fraction matrix
```{R}
# pre-process the matrix
t_alt_frac[is.na(t_alt_frac)] <- 0; dim(t_alt_frac) # fill NaNs with 0
t_alt_frac <- t_alt_frac[, colSums(t_alt_frac) > 0]; dim(t_alt_frac) # remove loci with all 0s
set.seed(42)
pca_result <- irlba::prcomp_irlba(
    t_alt_frac,
    n = 30,
    center = TRUE,
    scale. = FALSE
)
pca_projection <- pca_result$x
write.csv(
  x = cbind(
    data.frame("Barcode" = rownames(t_alt_frac)),
    pca_projection
  ),
  file = "outputs/notebooks/UMAP-show-scSNV-Egr/pca_projection_all_loci.csv",
  quote = FALSE,
  row.names = FALSE
)
```

UMAP on PCA projection
```{R}
n_neighbors <- 5
min_dist <- 0.1
metric <- "euclidean"
seed <- 42
umap_snv <- uwot::umap(
    pca_projection,
    n_neighbors = n_neighbors,
    min_dist = min_dist,
    spread = 1,
    metric = metric,
    n_components = 2,
    seed = seed,
    verbose = TRUE
)
umap_snv <- as.data.table(umap_snv)
setnames(umap_snv, "V1", "UMAP.1")
setnames(umap_snv, "V2", "UMAP.2")
umap_snv$Barcode <- rownames(t_alt_frac)
umap_snv[,c("Barcode", "UMAP.1", "UMAP.2")] %>% write.csv(
  "outputs/notebooks/UMAP-show-scSNV-Egr/snv-umap.csv",
  quote = FALSE,
  row.names = FALSE
)
```

Plot UMAP
SNV UMAP colored by Gene expression clusters
```{R}
umap_snv <- fread("outputs/notebooks/UMAP-show-scSNV-Egr/snv-umap.csv")
setkey(umap_snv, Barcode)
umap_snv_subset <- umap_snv[intersect(umap_snv$Barcode, cell_types$Barcode), ]
umap_snv_subset$Color_ge <- cell_types[umap_snv_subset$Barcode, "Color"]
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/SNV-UMAP_Tung-color.svg",
    width = 10,
    height = 10,
    bg = "white"
)
plot(
    x=umap_snv_subset[["UMAP.1"]],
    y=umap_snv_subset[["UMAP.2"]],
    col=umap_snv_subset[["Color_ge"]],
    pch=19,
    cex=0.5,
    main="UMAP of SNV data\ncolored by Tung Fig.1 cell type\nPCA-scaled=F",
    xlab="UMAP1",
    ylab="UMAP2"
)
dev.off()
```

Perform k-means clustering based on PCA projection
```{R}
# set k-means parameters
set.seed(42)
k <- 10
nstart <- 25
kmeans_res <- kmeans(
    pca_projection,
    centers = k,
    nstart = nstart
)
# re-label clusters based on sizes
sizes <- table(kmeans_res$cluster)
new_labels <- as.numeric(
    factor(
        kmeans_res$cluster,
        levels = names(sort(sizes, decreasing = TRUE))
    )
)
snv_kmeans <- data.frame(
    "Barcode" = barcodes[,1],
    "Cluster" = new_labels
)
snv_kmeans$Color <- color_snv[as.character(new_labels)]
rownames(snv_kmeans) <- snv_kmeans$Barcode
write.csv(snv_kmeans, file = "outputs/notebooks/UMAP-show-scSNV-Egr/snv-cluster.csv")
# added color to umap_snv
umap_snv$Color_snv <- snv_kmeans[rownames(umap_snv), "Color"]
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/SNV-UMAP_SNV-kmeans.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x = umap_snv[, "UMAP.1"],
    y = umap_snv[, "UMAP.2"],
    col = umap_snv[, "Color_snv"],
    pch = 19,
    cex = 0.5,
    main = "UMAP of SNV data\ncolored by k-means clustering\nPCA-scaled=F",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

Plot original UMAP colored by k-means clustering
```{R}
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/GE-UMAP_SNV-kmeans.svg",
    width = 7,
    height = 7,
    bg = "white"
)
umap_gene_expression$Color_snv <- snv_kmeans[umap_gene_expression$Barcode, "Color"]
plot(
    x = umap_gene_expression[,"UMAP.1"],
    y = umap_gene_expression[,"UMAP.2"],
    col = umap_gene_expression$Color_snv,
    pch = 19,
    cex = 0.5,
    main = "UMAP of gene expression\ncolored by SNV k-means clusters",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

Plot original UMAP colored by Tung Fig. 1 clustering

```{R}
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/GE-UMAP_Tung-color.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x = umap_gene_expression[,"UMAP.1"],
    y = umap_gene_expression[,"UMAP.2"],
    col = umap_gene_expression$Color_ge,
    pch = 19,
    cex = 0.5,
    main = "UMAP of gene expression\ncolored by SNV k-means clusters",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

# Run UMAP-PCA on alternate fraction matrix (using only homozygous loci)
```{R}
# pre-process the matrix
alt_frac_homo <- alt_frac[rownames(hom_ref), ]
t_alt_frac_homo <- t(alt_frac_homo)
t_alt_frac_homo[is.na(t_alt_frac_homo)] <- 0; dim(t_alt_frac_homo) # fill NaNs with 0
t_alt_frac_homo <- t_alt_frac_homo[, colSums(t_alt_frac_homo) > 0]; dim(t_alt_frac_homo) # remove loci with all 0s
set.seed(42)
pca_result_homo <- irlba::prcomp_irlba(
    t_alt_frac_homo,
    n = 30,
    center = TRUE,
    scale. = FALSE
)
pca_projection_homo <- pca_result_homo$x
write.csv(
  x = cbind(
    data.frame("Barcode" = rownames(t_alt_frac_homo)),
    pca_projection_homo
  ),
  file = "outputs/notebooks/UMAP-show-scSNV-Egr/pca_projection_homo_loci.csv",
  quote = FALSE,
  row.names = FALSE
)
```

UMAP on PCA projection
```{R}
n_neighbors <- 15
min_dist <- 0.1
metric <- "euclidean"
seed <- 42
umap_snv_homo <- uwot::umap(
    pca_projection_homo,
    n_neighbors = n_neighbors,
    min_dist = min_dist,
    spread = 1,
    metric = metric,
    n_components = 2,
    seed = seed,
    verbose = TRUE
)
umap_snv_homo <- as.data.table(umap_snv_homo)
setnames(umap_snv_homo, "V1", "UMAP.1")
setnames(umap_snv_homo, "V2", "UMAP.2")
umap_snv_homo$Barcode <- rownames(t_alt_frac_homo)
umap_snv_homo[,c("Barcode", "UMAP.1", "UMAP.2")] %>% write.csv(
  "outputs/notebooks/UMAP-show-scSNV-Egr/snv-umap_homo.csv",
  quote = FALSE,
  row.names = FALSE
)
setkey(umap_snv_homo, Barcode)
```
Plot UMAP
SNV UMAP colored by Gene expression clusters
```{R}
umap_snv_homo_subset <- umap_snv_homo[intersect(umap_snv_homo$Barcode, cell_types$Barcode), ]
umap_snv_homo_subset$Color_ge <- cell_types[umap_snv_homo_subset$Barcode, "Color"]
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/HOMO_SNV-UMAP_Tung-color.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x=umap_snv_homo_subset[["UMAP.1"]],
    y=umap_snv_homo_subset[["UMAP.2"]],
    col=umap_snv_homo_subset[["Color_ge"]],
    pch=19,
    cex=0.5,
    main="UMAP of SNV data (homo)\ncolored by Tung Fig.1 cell type\nPCA-scaled=F",
    xlab="UMAP1",
    ylab="UMAP2"
)
dev.off()
```

Perform k-means clustering based on PCA projection

```{R}
# set k-means parameters
set.seed(42)
k <- 10
nstart <- 25
kmeans_res <- kmeans(
    pca_projection,
    centers = k,
    nstart = nstart
)
# re-label clusters based on sizes
sizes <- table(kmeans_res$cluster)
new_labels <- as.numeric(
    factor(
        kmeans_res$cluster,
        levels = names(sort(sizes, decreasing = TRUE))
    )
)
snv_kmeans <- data.frame(
    "Barcode" = barcodes[,1],
    "Cluster" = new_labels
)
snv_kmeans$Color <- color_snv[as.character(new_labels)]
rownames(snv_kmeans) <- snv_kmeans$Barcode
write.csv(snv_kmeans, file = "outputs/notebooks/UMAP-show-scSNV-Egr/snv-cluster.csv")
# added color to umap_snv
umap_snv$Color_snv <- snv_kmeans[rownames(umap_snv), "Color"]
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/SNV-UMAP_SNV-kmeans.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x = umap_snv[, "UMAP.1"],
    y = umap_snv[, "UMAP.2"],
    col = umap_snv[, "Color_snv"],
    pch = 19,
    cex = 0.5,
    main = "UMAP of SNV data\ncolored by k-means clustering\nPCA-scaled=F",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

Plot original UMAP colored by k-means clustering

```{R}
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/GE-UMAP_SNV-kmeans.svg",
    width = 7,
    height = 7,
    bg = "white"
)
umap_gene_expression$Color_snv <- snv_kmeans[umap_gene_expression$Barcode, "Color"]
plot(
    x = umap_gene_expression[,"UMAP.1"],
    y = umap_gene_expression[,"UMAP.2"],
    col = umap_gene_expression$Color_snv,
    pch = 19,
    cex = 0.5,
    main = "UMAP of gene expression\ncolored by SNV k-means clusters",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

Plot original UMAP colored by Tung Fig. 1 clustering

```{R}
# plot UMAP
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/GE-UMAP_Tung-color.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x = umap_gene_expression[,"UMAP.1"],
    y = umap_gene_expression[,"UMAP.2"],
    col = umap_gene_expression$Color_ge,
    pch = 19,
    cex = 0.5,
    main = "UMAP of gene expression\ncolored by SNV k-means clusters",
    xlab = "UMAP1",
    ylab = "UMAP2"
)
dev.off()
```

# Run UMAP-PCA on alternate fraction matrix (using het + snp loci)
```{R}
# pre-process the matrix
alt_frac_het <- alt_frac[c(rownames(het), rownames(snp)), ]
t_alt_frac_het <- t(alt_frac_het)
t_alt_frac_het[is.na(t_alt_frac_het)] <- 0; dim(t_alt_frac_het) # fill NaNs with 0
t_alt_frac_het <- t_alt_frac_het[, colSums(t_alt_frac_het) > 0]; dim(t_alt_frac_het) # remove loci with all 0s
set.seed(42)
pca_result_het <- irlba::prcomp_irlba(
    t_alt_frac_het,
    n = 30,
    center = TRUE,
    scale. = FALSE
)
pca_projection_het <- pca_result_het$x
write.csv(
  x = cbind(
    data.frame("Barcode" = rownames(t_alt_frac_het)),
    pca_projection_het
  ),
  file = "outputs/notebooks/UMAP-show-scSNV-Egr/pca_projection_het_loci.csv",
  quote = FALSE,
  row.names = FALSE
)
```

UMAP on PCA projection
```{R}
n_neighbors <- 15
min_dist <- 0.1
metric <- "euclidean"
seed <- 42
umap_snv_het <- uwot::umap(
    pca_projection_het,
    n_neighbors = n_neighbors,
    min_dist = min_dist,
    spread = 1,
    metric = metric,
    n_components = 2,
    seed = seed,
    verbose = TRUE
)
umap_snv_het <- as.data.table(umap_snv_het)
setnames(umap_snv_het, "V1", "UMAP.1")
setnames(umap_snv_het, "V2", "UMAP.2")
umap_snv_het$Barcode <- rownames(t_alt_frac_het)
umap_snv_het[,c("Barcode", "UMAP.1", "UMAP.2")] %>% write.csv(
  "outputs/notebooks/UMAP-show-scSNV-Egr/snv-umap_het.csv",
  quote = FALSE,
  row.names = FALSE
)
setkey(umap_snv_homo, Barcode)
```
Plot UMAP
SNV UMAP colored by Gene expression clusters
```{R}
umap_snv_het_subset <- umap_snv_het[intersect(cell_types$Barcode, umap_snv_het$Barcode), ]
umap_snv_het_subset$Color_ge <- cell_types[umap_snv_het_subset$Barcode, "Color"]
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/HET_SNV-UMAP_Tung-color.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x=umap_snv_het_subset[["UMAP.1"]],
    y=umap_snv_het_subset[["UMAP.2"]],
    col=umap_snv_het_subset[["Color_ge"]],
    pch=19,
    cex=0.5,
    main="UMAP of SNV data (het)\ncolored by Tung Fig.1 cell type\nPCA-scaled=F",
    xlab="UMAP1",
    ylab="UMAP2"
)
dev.off()
```

# EDA on DNA variant calling results
```{R}
dim(t_alt_frac) # [1]   4870 187320
nrow(hom_alt) # [1] 413
nrow(hom_ref) # [1] 30470
nrow(het) # [1] 59577
nrow(snp) # [1] 137855
loci_hom_alt <- rownames(hom_alt)
loci_hom_ref <- rownames(hom_ref)
loci_het <- rownames(het)
loci_snp <- rownames(snp)
(loci_snp_het <- intersect(loci_snp, loci_het)) %>% length() # [1] 59075
(loci_snp_hom_ref <- intersect(loci_snp, loci_hom_ref)) %>% length() # [1] 0
(loci_snp_hom_alt <- intersect(loci_snp, loci_hom_alt)) %>% length() # [1] 407
(loci_snp_filtered <- loci_snp %>%
    setdiff(., loci_het) %>%
    setdiff(., loci_hom_ref) %>%
    setdiff(., loci_hom_alt)) %>% length() # [1] 78373
snp_het <- snp[loci_snp_het,]
snp_hom_alt <- snp[loci_snp_hom_alt,]
snp_filtered <- snp[loci_snp_filtered,]
```

See GATK SNP sites that are identified as both SNP and heterozygous no-variant sites

```{R}
svg("outputs/notebooks/UMAP-show-scSNV-Egr/GATK-SNP-type.svg")
table(apply(snp[,c("REF", "ALT")], 1, paste, collapse = "->")) %>%
    as.data.frame %>%
    subset(Freq > 50) %>%
    arrange(desc(Freq)) %>%
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "GATK SNP sites (plot types where #sites > 50)",
        x = "Type of SNP",
        y = "#sites"
    )
dev.off()

svg("outputs/notebooks/UMAP-show-scSNV-Egr/GATK-filtered-SNP-type_F_PASS=0.2.svg")
table(apply(snp_filtered[,c("REF", "ALT")], 1, paste, collapse = "->")) %>%
    as.data.frame %>%
    subset(Freq > 50) %>%
    arrange(desc(Freq)) %>%
    ggplot(aes(x = Var1, y = Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "Filtered GATK SNP sites (plot types where #sites > 50)",
        x = "Type of SNP",
        y = "#sites"
    )
dev.off()
```

See scRNA editing sites

```{R}
svg("outputs/notebooks/UMAP-show-scSNV-Egr/scRNA-editing-type_F_PASS=0.2.svg")
type_counts <- sc_vcf[loci_hom_ref, c("REF", "ALT")] %>% 
    apply(1, paste, collapse = "->") %>%
    table %>%
    as.data.frame
colnames(type_counts) <- c("Type", "Freq")
type_counts %>%
    subset(Freq > 50) %>%
    arrange(desc(Freq)) %>%
    ggplot(aes(x = Type, y = Freq)) +
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
        title = "scRNA editing sites (hom-ref; F_PASS=0.2; plot types where #sites > 50)",
        x = "Type of editing",
        y = "#sites"
    )
dev.off()
```

Histogram of alternate fraction at homozygous loci with high variance

```{R}
# plot histogram of alternate fraction at homozygous loci
locus_stats_hom_ref <- locus_stats[loci_hom_ref,]
locus_stats_hom_ref <- locus_stats_hom_ref[order(locus_stats_hom_ref$locus_n_cells, decreasing=TRUE),]
for (i in 1:10) {
    locus <- rownames(locus_stats_hom_ref)[i]
    svg(
        sprintf(
            "outputs/notebooks/UMAP-show-scSNV-Egr/alt_frac_hist_%s_hom_ref.svg",
            locus
        ),
        width = 7,
        height = 7,
        bg = "transparent"
    )
    alt_frac_locus <- t_alt_frac[, locus] %>% 
        unlist %>%
        hist(
            main = sprintf(
                "Alternate fraction at locus %s\n%s cells; variance=%s",
                locus,
                sum(!is.na(t_alt_frac[, locus])),
                locus_stats[locus, "locus_var"]
            ),
            xlab = "Alternate fraction",
            ylab = "Observed in N cells",
            breaks = 100
        )
    dev.off()
}
```

Count high-quality samples supporting each R\|R sites using hom_ref.tsv

```{R}
library(ggplot2)

# 
RGQ_columns <- hom_ref_unfiltered_tsv[, grep("RGQ", colnames(hom_ref_unfiltered_tsv)), drop = FALSE]
RGQ_passed <- RGQ_columns > 20
RGQ_passed[is.na(RGQ_passed)] <- FALSE
DP_columns <- hom_ref_unfiltered_tsv[, grep("DP", colnames(hom_ref_unfiltered_tsv)), drop = FALSE]
DP_passed <- DP_columns > 10
DP_passed[is.na(DP_passed)] <- FALSE
ALL_passed <- RGQ_passed & DP_passed
num_passed <- apply(ALL_passed, 1, sum)

# make plot data
# Add marginal histograms
# ggMarginal(p, type = "histogram", fill = "steelblue", bins = 30)

svg("outputs/notebooks/UMAP-show-scSNV-Egr/genotyping/hist_#hq_supporting_R|R.svg")
hist(
    num_passed,
    breaks = 50,
    xlab = "No. high-confidence R|R sample\n(Genotype quality > 20 & Read depth > 10)",
    ylab = "Counts",
    main = "66,571 loci without high-confidence A|A or R|A sample"
)
lines(
    x = c(10,10),
    y = c(0,5000),
    lwd = 2,
    lty = 5,
    col = "red"
)
dev.off()
```

# LCM RNA editing comes in here

Correlation of RNA editing between replicates

```{R}
# plot correlation of RNA editing between replicates
corr_pairwise_filled_pearson <- matrix(
    nrow = length(path_lcm_editing),
    ncol = length(path_lcm_editing)
)
colnames(corr_pairwise_filled_pearson) <- names(path_lcm_editing)
rownames(corr_pairwise_filled_pearson) <- names(path_lcm_editing)
for (i in 1:length(path_lcm_editing)) {
    for (j in 1:length(path_lcm_editing)) {
        if (i != j) {
            x = lcm_editing_freq[, i]
            y = lcm_editing_freq[, j]
            # remove loci with NA in both replicates
            not_both_na = !is.na(x) & !is.na(y)
            x = x[not_both_na]
            y = y[not_both_na]
            # fill NA in one replicate with 0
            x[is.na(x)] <- 0
            y[is.na(y)] <- 0
            corr_pairwise_filled_pearson[i, j] <- cor(x, y, use = "everything", method = "pearson")
        } else {
            corr_pairwise_filled_pearson[i, j] <- 1
        }
    }
}
write.csv(
    corr_pairwise_filled_pearson,
    "outputs/notebooks/UMAP-show-scSNV-Egr/lcm-editing-correlation-pearson-prevent-0-inflation.csv"
)
write.csv(
    cor(lcm_editing_freq, use = "pairwise.complete.obs", method = "pearson"),
    "outputs/notebooks/UMAP-show-scSNV-Egr/lcm-editing-correlation-pearson-pairwise.complete.obs.csv"
)
```

Correlation between LCM RNA editing and scRNA editing (1) Filled NA in scRNA editing with 0

```{R}
# fill alt_frac with 0
alt_frac_filled <- alt_frac
alt_frac_filled[is.na(alt_frac_filled)] <- 0
alt_frac_filled <- as.matrix(alt_frac_filled)
```

(2) Visualize correlation between LCM RNA editing and scRNA editing on UMAP embedding

```{R}
# prevent re-running this
alt_frac_mat <- as.matrix(alt_frac)
```

```{R}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
# 
params1 <- list(method="pearson", filled=TRUE, umap="GE")
params2 <- list(method="pearson", filled=TRUE, umap="SNV")
params3 <- list(method="pearson", filled=FALSE, umap="GE")
params4 <- list(method="pearson", filled=FALSE, umap="SNV")
# 
params5 <- list(method="spearman", filled=TRUE, umap="GE")
params6 <- list(method="spearman", filled=TRUE, umap="SNV")
params7 <- list(method="spearman", filled=FALSE, umap="GE")
params8 <- list(method="spearman", filled=FALSE, umap="SNV")

params <- params3

# parse variables
method <- params$method
filled <- params$filled
umap <- params$umap
if (filled) {
    x <- alt_frac_filled
    y <- lcm_editing_freq_filled
    use <- "everything"
} else {
    x <- alt_frac_mat
    y <- lcm_editing_freq
    use <- "pairwise.complete.obs"
}
fout <- sprintf(
    "outputs/notebooks/UMAP-show-scSNV-Egr/LCM_correlation/%s-UMAP_LCM_Correlation_%s_Fill_%s.svg",
    umap,
    method,
    filled
)
if(umap == "GE") {
    umap_df <- umap_gene_expression[, c("Barcode", "UMAP.1", "UMAP.2")]    
} else if (umap == "SNV") {
    umap_df <- umap_snv %>% rownames_to_column("Barcode")
} else {
    stop()
}

# calculate correlation between each cell and each LCM sample
rm(corr_cell_lcm)
corr_cell_lcm <- cor(
    x = x,
    y = y,
    use = use,
    method = method
)

# Convert correlation matrix to long format
corr_long <- corr_cell_lcm %>%
    as.data.frame() %>%
    rownames_to_column("Barcode") %>%
    pivot_longer(-Barcode, names_to = "LCM_Sample", values_to = "Correlation")

# Filter to barcodes present in UMAP
corr_filtered <- corr_long %>%
    filter(Barcode %in% umap_df$Barcode)

# Join UMAP coordinates with correlation values
plot_data <- left_join(umap_df, corr_filtered, by = "Barcode")

# Create ggplot object
p <- ggplot(plot_data, aes(x = UMAP.1, y = UMAP.2, color = Correlation)) +
    geom_point(size = 0.2) +
    facet_wrap(~LCM_Sample, ncol = 3) +
    scale_colour_gradient2(
        low = "blue",
        mid = "white",
        high = "red",
        midpoint = 0,
        name = "Correlation"
    ) +
    theme_minimal() +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()
    ) +
    labs(
        x = "UMAP1",
        y = "UMAP2",
        title = sprintf("%s, %s, %s", umap, method, use)
    )

# Save as SVG
svg(fout, width = 10, height = 8, bg = "white")
print(p)
dev.off()
```

# TODO: Intersect between LCM RNA editing and scRNA editing

```{R}
# find intersection between LCM RNA editing and scRNA editing
overlap_cell_lcm <- matrix(
    nrow = ncol(alt_frac),
    ncol = ncol(lcm_editing_freq)
)
for (i in 1:ncol(alt_frac)) {
    for (j in 1:ncol(lcm_editing_freq)) {
        overlap_cell_lcm[i, j] <- length(
            intersect(
                rownames(alt_frac)[!is.na(alt_frac[, i])],
                rownames(lcm_editing_freq)[!is.na(lcm_editing_freq[, j])]
            )
        )
    }
}
```

ChatGPT says this is faster:

```{R}
# Get non-NA row names for each column once
alt_frac_list <- apply(alt_frac, 2, function(col) rownames(alt_frac)[!is.na(col)])
lcm_editing_list <- apply(lcm_editing_freq, 2, function(col) rownames(lcm_editing_freq)[!is.na(col)])

# Create a function to compute intersection length
intersect_length <- function(a, b) length(intersect(a, b))

# Vectorize the function to use with outer
vec_intersect_length <- Vectorize(intersect_length)

# Compute the matrix
overlap_cell_lcm <- outer(alt_frac_list, lcm_editing_list, vec_intersect_length)
```

# TODO: plot correlation of LCM RNA editing with scRNA editing on hom-ref, not-hom-ref loci

````{=html}
<!-- # EDA on DNA SNV
NOTE: go to the second top chunk if `dna_snv` is not loaded
```{R}
# plot histogram of DNA SNV (including AF == 1)
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/dna_snv_hist.svg",
    width = 7,
    height = 7,
    bg = "white"
)
hist(
    dna_snv$AF,
    main = "DNA SNV alternate fraction",
    xlab = "Alternate fraction",
    ylab = "No. loci",
    breaks = 100
)
dev.off()
# plot histogram of DNA SNV (excluding AF == 1)
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/dna_snv_hist_exclude_AF=1.svg",
    width = 7,
    height = 7,
    bg = "white"
)
hist(
    dna_snv$AF[dna_snv$AF < 1],
    main = "DNA SNV alternate fraction\nexcluding AF=1",
    xlab = "Alternate fraction",
    ylab = "No. loci",
    breaks = 100
)
dev.off()
# plot scatter plot of AF v.s. DP (to see if AF is correlated with DP)
# Note: exclude AF == 1 loci
svg(
    "outputs/notebooks/UMAP-show-scSNV-Egr/dna_snv_AF~DP.svg",
    width = 7,
    height = 7,
    bg = "white"
)
plot(
    x = dna_snv$DP[dna_snv$AF < 1],
    y = dna_snv$AF[dna_snv$AF < 1],
    main = "DNA SNV alternate fraction v.s. depth of coverage\nexcluding AF=1",
    xlab = "Depth of coverage",
    ylab = "Alternate fraction",
    cex = 0.1
)
abline(lm(dna_snv$AF[dna_snv$AF < 1] ~ dna_snv$DP[dna_snv$AF < 1]), col = "red")
dev.off()
``` -->
````
